---
title: "Beluga ddRADseq Inbreeding Coefficient Analyses"
author: "Audrey Salinger"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(vcfR)
library(readr)
library(reshape2)
library(inbreedR)
library(cowplot)
library(gridGraphics)
library(ggplot2)
library(arsenal)
require(brms)
require(posterior)
require(bayesplot)
require(ggridges)
require(stringr)

setwd("/Users/audreysalinger/PhD/beluga/r_analyses")
```

## Objective

I have completed bioinformatic analysis of the ddRAD sequencing data for the beluga project using two alpha value thresholds for calling SNPs and calling genotypes: 0.01 and 0.001.  
  
Summary of the total SNPs passing filters for the two data sets:  
**0.01  = 28,852 (154 samples)**  
**0.001 = 16,122 (154 samples)**  

Moving forward, I will focus on/present values for the more stringent data set, 0.001, since it allows for higher confidence in the resulting SNPs and genotype calls. However, I will still look at the 0.01 data set here to understand the data.

*In this file I will detail the process used for the remaining analyses, including estimating inbreeding coefficients, generating plots, and performing the Bayesian statistical analysis.*

First I will import the final VCFs:
```{r import vcfs}
vcf_0.01 = read.vcfR("final-vcfs/0.01_fully_filtered.vcf.gz", verbose = FALSE)
vcf_0.001 = read.vcfR("final-vcfs/0.001_fully_filtered.vcf.gz", verbose = FALSE)
```

## 1. Estimating inbreeding coefficients

I want to calculate inbreeding coefficients for each individual. There are many different methods to estimate individual inbreeding coefficients; I will calculate F, sMLH, IR, and HL.

### Calculating F with VCFtools

I will first use `VCFtools --het` to calculate the inbreeding coefficient F for each individual; the remainder of the analysis will use R.  
  
The resulting value F represents the deviation between observed and expected heterozygosity within an individual.

```{r import vcftools het}
F_0.01 = read_tsv("inbreeding-coefficients/0.01_vcftools.het", col_names = TRUE)
F_0.001 = read_tsv("inbreeding-coefficients/0.001_vcftools.het", col_names = TRUE)
```

### Calculate sMLH with InbreedR

**Standardized multi-locus heterozygosity (sMLH)** is defined by InbreedR as the total number of heterozygous loci in an individual divided by the sum of average observed heterozygosities in the population over the subset of loci successfully typed in the focal individual. This inbreeding coefficient is a way to assess heterozygosity across an indivdual's genome relative to the heterozygosity observed in the population/sample set. Because individuals are typed at different numbers of marker loci, standardizing ensures that the heterozygosity of all individuals is measured on the same scale.  
  
*From Coltman, D. W., Pilkington, J. G., Smith, J. A., & Pemberton, J. M. (1999). Parasite-mediated selection against inbred Soay sheep in a free-living, island population. Evolution, 1259-1267.*  

#### Convert VCF to InbreedR working format

For InbreedR, the input needs to be an individual x locus matrix or data.frame in which rows represent individuals and each column represents a locus. If an individual is heterozygous at a given locus, it is coded as 1, whereas a homozygote is coded as 0,and missing data are coded as NA.  
  
First, I can use vcfR `extract.gt` to generate a matrix of loci x individuals. I can then transpose the matrix to get the proper individual x locus format.  
  
InbreedR has a function `convert_raw` that converts raw genotype data into the required working format. This conversion function requires a raw genotype matrix with individuals in rows and each locus in two adjacent columns.  

I will need to split my genotypes (which are in format 0/1) into two adjacent columns, one for each allele.  
  
Then I can use the `convert_raw` function for the final conversion step. Inbreed R also has a function `check_data` that will make sure the data are in the correct formatting; it checks that: (1) the genotype data just contains 3 elements, which is 0 for homozygote, 1 for heterozygote and NA for missing data, (2) the number of individuals corresponds to the number of rows and the number of loci corresponds to the number of columns, (3) the data type is numeric.  

```{r InbreedR data setup}
# extract genotypes and convert to allelic state 0/1
# output is a matrix of genotypes w/ a column for each individual & a row for each SNP/locus
genotypes_01 = extract.gt(vcf_0.01)
genotypes_001 = extract.gt(vcf_0.001)

# transpose the matrix so that rows are individuals & and each column is a locus
genotypes_01 = t(genotypes_01)
genotypes_001 = t(genotypes_001)

# split genotypes into two adjacent columns, one for each allele 
# apply(matrix, 2 = apply to columns, apply the function colsplit)
# colsplit(x = string to split up, / = expression to split on, a & b = suffixes of new columns)
# cbind to re-bind together all the individual locus columns that have been split in 2
split_gt_01 = do.call(cbind, apply(genotypes_01, 2, function(x) colsplit(x, "/", c("a","b"))))
split_gt_001 = do.call(cbind, apply(genotypes_001, 2, function(x) colsplit(x, "/", c("a","b"))))

# convert data to inbreedR input format
inbreedR_input_01 = inbreedR::convert_raw(split_gt_01)
inbreedR_input_001 = inbreedR::convert_raw(split_gt_001)

# check the data formatting
check_data(inbreedR_input_01, num_ind = 154, num_loci = 28852)
check_data(inbreedR_input_001, num_ind = 154, num_loci = 16122)
```

In order to check that this conversion has maintained data integrity, I will ensure that the first 5 and last 5 loci for the first 5 and last 5 samples are still in agreement for both data sets.

First, extracting this data from the original VCFs:
```
#!/bin/bash

#################################################################################
#  extract first 5 and last 5 genotypes for the first 5 and last 5 individuals  #
#                  to check against transformed genotype files                  #
#################################################################################

cd ../final-vcfs

###################
#  0.01 data set  #
###################

#--- get first 5 sample names and save to a temp file---#
zgrep -v "^##" 0.01_fully_filtered.vcf.gz | head -n 1 | cut -f 10-14 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,8)}1' > 0.01_first5ind

#--- add first 5 genotypes for the first 5 samples to temp file---#
zgrep -v "^##" 0.01_fully_filtered.vcf.gz | head -n 6 | tail -n 5 | cut -f 10-14 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,3)}1' >> 0.01_first5ind

#--- add last 5 genotypes for the first 5 samples to temp file ---#
zgrep -v "^##" 0.01_fully_filtered.vcf.gz | tail -n 5 | cut -f 10-14 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,3)}1' >> 0.01_first5ind


#--- get last 5 sample names and save to another temp file ---#
zgrep -v "^##" 0.01_fully_filtered.vcf.gz | head -n 1 | cut -f 159-163 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,8)}1' > 0.01_last5ind

#--- add first 5 genotypes for the last 5 samples to temp file ---#
zgrep -v "^##" 0.01_fully_filtered.vcf.gz | head -n 6 | tail -n 5 | cut -f 159-163 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,3)}1' >> 0.01_last5ind

#--- add last 5 genotypes for the last 5 samples to temp file ---#
zgrep -v "^##" 0.01_fully_filtered.vcf.gz | tail -n 5 | cut -f 159-163 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,3)}1' >> 0.01_last5ind


#--- paste temp files together ---#
paste -d ' ' 0.01_first5ind 0.01_last5ind > ../gt-conversion-check/0.01_vcf_gt_subset

#--- remove temp files ---#
rm 0.01_first5ind
rm 0.01_last5ind


####################
#  0.001 data set  #
####################

#--- get first 5 sample names and save to a temp file---#
zgrep -v "^##" 0.001_fully_filtered.vcf.gz | head -n 1 | cut -f 10-14 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,8)}1' > 0.001_first5ind

#--- add first 5 genotypes for the first 5 samples to temp file---#
zgrep -v "^##" 0.001_fully_filtered.vcf.gz | head -n 6 | tail -n 5 | cut -f 10-14 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,3)}1' >> 0.001_first5ind

#--- add last 5 genotypes for the first 5 samples to temp file ---#
zgrep -v "^##" 0.001_fully_filtered.vcf.gz | tail -n 5 | cut -f 10-14 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,3)}1' >> 0.001_first5ind


#--- get last 5 sample names and save to another temp file ---#
zgrep -v "^##" 0.001_fully_filtered.vcf.gz | head -n 1 | cut -f 159-163 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,8)}1' > 0.001_last5ind

#--- add first 5 genotypes for the last 5 samples to temp file ---#
zgrep -v "^##" 0.001_fully_filtered.vcf.gz | head -n 6 | tail -n 5 | cut -f 159-163 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,3)}1' >> 0.001_last5ind

#--- add last 5 genotypes for the last 5 samples to temp file ---#
zgrep -v "^##" 0.001_fully_filtered.vcf.gz | tail -n 5 | cut -f 159-163 | awk '{for(i=1;i<=NF;i++) $i=substr($i,1,3)}1' >> 0.001_last5ind


#--- paste temp files together ---#
paste -d ' ' 0.001_first5ind 0.001_last5ind > ../gt-conversion-check/0.001_vcf_gt_subset

#--- remove temp files ---#
rm 0.001_first5ind
rm 0.001_last5ind
```

```{r check genotype conversion integrity}
# import gt subsets from original vcfs
gt_check_01 = read_delim("gt-conversion-check/0.01_vcf_gt_subset", delim = ' ', col_names = TRUE)
gt_check_001 = read_delim("gt-conversion-check/0.001_vcf_gt_subset", delim = ' ', col_names = TRUE)

# extract the same subset from the genotypes matrices created in InbreedR data setup (and edit to match formatting)
genotypes_01_subset = as.data.frame(t(genotypes_01[c(1:5, 150:154), c(1:5, 28848:28852)]))
genotypes_01_subset[is.na(genotypes_01_subset)] = './.'

genotypes_001_subset = as.data.frame(t(genotypes_001[c(1:5, 150:154), c(1:5, 16118:16122)]))
genotypes_001_subset[is.na(genotypes_001_subset)] = './.'

# compare genotype matrix to the original vcf
gt_check_01 == genotypes_01_subset
gt_check_001 == genotypes_001_subset
```

For this subset, the genotypes matrices created do match the original vcfs, as expected. I will check further by comparing this subset of genotypes from the original vcfs to the final InbreedR input data frames (where columns are loci, 0 = homozygous, and 1 = heterozygous).

```{r subset InbreedR inputs for conversion check}
as.data.frame(t(inbreedR_input_01[c(1:5, 150:154), c(1:5, 28848:28852)]))
as.data.frame(t(inbreedR_input_001[c(1:5, 150:154), c(1:5, 16118:16122)]))
```

Indeed for the subset, the original vcf genotypes and the genotypes in the final InbreedR converted files match. All good to proceed!

#### Identity disequilibrium

Before moving on with the calculation of inbreeding coefficients, I first want to estimate identity disequilibrium (ID, g2) for the data using InbreedR's `g2_snps` function.  
  
Identity disquilibrium is the phenomenon in which variation in inbreeding causes heterozygosity to be correlated across loci. In order to (theoretically) detect inbreeding depression, markers should have nonzero variance and heterozygosity of markers should correlate with the heterozygosity of functional loci. Assessing identity disequilibrium will assess the potential of these data to detect heterozygosity-fitness correlations (HFCs).  
  
g2 is a measure of identity disequilibrium, which reflects how heterozygosity (or homozygosity) is correlated across markers/loci. The `g2_snps` function in InbreedR permutes the genetic data to generate a p-value for the null hypothesis of no variance in inbreeding in the samples (i.e., g2 = 0, all individuals experience the same level of inbreeding). Low g2 values may suggest low expected correlations between heterozygosity and inbreeding.

```{r InbreedR g2 - Figure S5, eval=FALSE, include=FALSE}
# for the 0.01 data set
g2_01 = g2_snps(inbreedR_input_01,
                nperm = 100, nboot = 10000, boot_over = "inds", CI = 0.95,
                parallel = TRUE, ncores = 4)

# for the 0.001 data set
g2_001 = g2_snps(inbreedR_input_001,
                 nperm = 100, nboot = 10000, boot_over = "inds", CI = 0.95,
                 parallel = TRUE, ncores = 4)

# save the results in a table
g2_100x10000 = data.frame(alpha = 0.001,
                          g2 = g2_001$g2,
                          CI_2.5 = g2_001$CI_boot[1],
                          CI_97.5 = g2_001$CI_boot[2],
                          p_val = g2_001$p_val,
                          nperm = 100,
                          nboot = 10000)

write.table(g2_100x10000, file = "g2_100x10000.tsv", sep = "\t", row.names = FALSE)

# plot the g2 outputs: the distribution of g2 estimates from bootstrap samples over individuals w/ 95% CIs
# black dot = the empirical g2 estimate

## open pdf file
pdf("plots/both_datasets/g2-plots.pdf", width = 7, height = 4)
## create plots
par(mfrow = c(1,2))
plot(g2_01, main = "0.01", col = "#88A2B9")
plot(g2_001, main = "0.001", col = "#88A2B9")
## close the file
dev.off()

## open pdf file
pdf("plots/g2-plot-0.001.pdf", width = 5, height = 4)
## create plots
par(mfrow = c(1,1))
plot(g2_001, col = "#88A2B9")
## close the file
dev.off()

## open pdf file
pdf("../supplementary_figures/Figure_S5.pdf", width = 5, height = 4)
## create plots
par(mfrow = c(1,1))
plot(g2_001, col = "#88A2B9", main = NA)
## close the file
dev.off()
```

A significant p-value indicates that we should reject the null hypothesis of no variance in inbreeding in the sample, therefore, for the 0.001 data set (which is the more stringent, conservative data set that I will be using in the final analysis) I can conclude that variance in inbreeding among individuals in the study was detected. I cannot draw this conclusion for the 0.01 data set.

Both data sets have g2 < 0, which suggests that heterozygosity at one locus is less likely occur with heterozygosity at another locus than expected by chance. So there may still be correlation or association between loci, but not in a way likely to be due to inbreeding.

#### Calculate sMLH

```{r InbreedR sMLH}
# calculate sMLH for both data sets
sMLH_0.01 = sMLH(inbreedR_input_01)
sMLH_0.001 = sMLH(inbreedR_input_001)

# create a data frame to hold the sMLH data; in the first column, shorten sample names to just Bel numbers
sMLH_df = data.frame(sample = substr(rownames(genotypes_01), 1, 8),
                     sMLH_0.01,
                     sMLH_0.001)

# save sMLH values to a file
write.table(sMLH_df, file = "inbreeding-coefficients/sMLH.tsv", sep = "\t", row.names = FALSE)
```

### Calculate IR and HL with GENHET

Next I want to calculate internal relatedness. Internal relatedness (IR) is a measure of of homozygosity that incorporates allele frequency, where being homozygous for a rarer allele will suggest a higher inbreeding estimate than being homozygous for a common allele. The equation, from Amos et al. 2001, is as follows:  
  
IR = (2H - ∑fi) / (2N - ∑fi)  
  
where H = number of homozygous loci, N = number of loci total, fi = frequency of the ith allele contained in the genotype.  
  
IR values centred more or less on zero suggest individuals born to "unrelated" parents. Negative values suggest relatively "outbred" individuals. High positive values are suggestive of inbreeding.  
  
I will also use GENHET to calculate homozygosity by locus (HL), which is a measure of homozygosity that weighs the contribution of each locus to the homozygosity index depending on their allelic variability in order to avoid underestimating heterozygosity of individuals carrying rare alleles. The equation, from Aparicio et al. 2006, is as follows:  
  
HL = ∑Eh / (∑Eh + ∑Ej)  
  
where Eh and Ej are the expected heterozygosities of the loci that an individual bears in homozygosis (h) and heterozygosis (j), respectively. 
  
GENHET is a function written for the program R that calculates 5 different individual heterozygosity estimates:  
1. proportion of heterozygous loci (PHt) in an individual: PHt = number of heterozygous loci / number of genotyped loci  
2. standardized heterozygosity based on the mean expected heterozygosity Hs_exp, Coltman 1999): Hs_exp = PHt / mean expected heterozygosity of typed loci  
3. standardized heterozygosity based on the mean observed heterozygosity (Hs_obs): Hs_obs = PHt / mean observed heterozygosity of typed loci  
4. internal relatedness (IR) (Amos 2001): IR = (2H – ∑fi) / (2N – ∑fi), where H is the number of loci that are homozygous, N is the number of loci and fi is the frequency of the ith allele contained in the genotype   
5. homozygosity by locus (HL) (Aparicio 2006): HL = ∑Eh / (∑Eh + ∑Ej), where Eh and Ej are the expected heterozygosities of the loci that an individual bears in homozygosis (h) and in heterozygosis (j), respectively.  
  
To use GENHET to calculate IR and HL, first need to create a file containing the genotypes of each individual:  
- the first column should contain the individuals’ identifiers (sample names)  
- the following columns contain the genotypes: 2 columns per locus: 1 for each allele; alleles have to be coded as numeric values; missing data must be coded as “NA”  
- the first row may contain the names of the columns, but that's not mandatory

```{r GENHET genotypes input}
# append a column of sample names to the existing split_gt data frames
genhet_input_01 = add_column(split_gt_01, sample = sMLH_df$sample, .before = 1)
genhet_input_001 = add_column(split_gt_001, sample = sMLH_df$sample, .before = 1)
```

To calculate individual heterozygosity estimates, GENHET needs allele frequencies. The manual for GENHET states that the simplest option is to ask GENHET to calculate the allele frequencies in your file of genotypes; to do this I must also provide a file giving the names of the loci in the same order as in the file of genotypes:
```{r GENHET loci input}
# create list of loci by capturing the columns of the input file less the first samples name column
loci_01 = colnames(genhet_input_01[-1])
loci_001 = colnames(genhet_input_001[-1])

# remove the trailing characters (e.g., :-.a), keep all characters through 5th from end, remove last 4 characters
loci_01 = str_sub(loci_01, end = -5)
loci_001 = str_sub(loci_001, end = -5)

# remove duplicate loci names (they each appear twice since that was the original input format)
loci_01 = unique(loci_01)
loci_001 = unique(loci_001)
```

I downloaded the GENHET function, manual, and example files directory into the working directory from http://www.aureliecoulon.net/research/ac-computer-programs.html.  
  
Load the GENHET code:
```{r load GENHET}
source("GENHETv3.1/GENHETv3.1.R")
```

Run GENHET:
```{r run GENHET for IR and HL, eval=FALSE, include=FALSE}
# to launch, must tell R what to name the output
# arguments: dat = genotypes object; estimfreq = calculate allele freq T (true) or F (false) , locname = loci list
genhet_0.01 = GENHET(dat = genhet_input_01, estimfreq = "T", locname = loci_01)
genhet_0.001 = GENHET(dat = genhet_input_001, estimfreq = "T", locname = loci_001)

# save outputs
write.table(genhet_0.01, file = "inbreeding-coefficients/GENHET_0.01_results.tsv", sep = "\t")
write.table(genhet_0.001, file = "inbreeding-coefficients/GENHET_0.001_results.tsv", sep = "\t")
```

GENHET takes quite a while to run, so I have commented out writing new files, to avoid damaging the saved versions.
For future analyses, I will read in the saved table.

```{r extract GENHET IR and HL}
# read in saved GENHET output tables
genhet_0.01_df = read.table("inbreeding-coefficients/GENHET_0.01_results.tsv")
genhet_0.001_df = read.table("inbreeding-coefficients/GENHET_0.001_results.tsv")

# extract IR and save IR values to their own data frame
IR_df = data.frame(sample = genhet_0.01_df$sampleid,
                   IR_0.01 = genhet_0.01_df$IR,
                   IR_0.001 = genhet_0.001_df$IR)

# extract HL and save HL values to their own data frame
HL_df = data.frame(sample = genhet_0.01_df$sampleid,
                   HL_0.01 = genhet_0.01_df$HL,
                   HL_0.001 = genhet_0.001_df$HL)

# save data frames
write.table(IR_df, file = "inbreeding-coefficients/IR.tsv", sep = "\t")
write.table(HL_df, file = "inbreeding-coefficients/HL.tsv", sep = "\t")
```

## 2. Plotting results

Import the mortality data:
```{r import and edit mortality data}
mortality = read.csv("beluga-mortalities.csv") %>% select(lab_code, date_found, sex, estimated_age, category, broad_category)

# filter mortality data set for just the individuals in the analysis
mortality = subset(mortality, lab_code %in% genetic_data$sample)

# create updated mortality classes
mortality = mortality %>% mutate(new_category = case_when(
            broad_category == "Dystocia / postpartum complication" ~ "Dystocia",
            broad_category == "Antropogenic causes" ~ "Anthropogenic causes",
            broad_category == "Neoplasia" ~ "Neoplasia",
            broad_category == "Infectious" ~ "Infectious disease",
            category == "Degenerative changes" ~ "Degenerative changes",
            category == "Undetermined" ~ "Undetermined",
            broad_category == "Neonatal mortality" ~ "Neonatal mortality"))

# create data frame for the live samples
live_samples = data.frame(lab_code = c("Bel00100", "Bel00739", "Bel01064", "Bel01664", "Bel00854"),
                          date_found = c(1995, 2009, 2014, 2022, 2000),
                          sex = c("M", "M", "M", "F", "M"),
                          estimated_age = NA,
                          category = "Live biopsy",
                          broad_category = "Live biopsy",
                          new_category = "Live biopsy")

# create a data frame that includes field info for both the dead and live samples
field_data = rbind(mortality, live_samples)
write.table(field_data, file = "field-data.tsv", sep = "\t", row.names = FALSE)

# number of males and females
field_data %>% dplyr::count(sex)

# date range
range(field_data$date_found)
```

Create a data frame with both field and genetic data:
```{r mortality plot data}
# create a data frame with sample name, cause of mortality, and inbreeding coefficients
combined_data = genetic_data %>% inner_join(field_data, by = join_by(sample == lab_code))

# reorder the new category factor to put live biopsy last
combined_data$new_category = factor(combined_data$new_category, levels = c("Anthropogenic causes", "Degenerative changes", "Dystocia", "Infectious disease", "Neonatal mortality",  "Neoplasia", "Undetermined", "Live biopsy"))
```

### Summary statistics for coefficients
```{r calculate coefficient summary stats}
# summarize all whales together 
coeff_summary_0.001 = data.frame(F = unclass(summary(F_0.001$F)),
                                 sMLH = unclass(summary(sMLH_df$sMLH_0.001)),
                                 IR = unclass(summary(IR_df$IR_0.001)),
                                 HL = unclass(summary(HL_df$HL_0.001)))
# add sd
coeff_summary_0.001[nrow(coeff_summary_0.001) + 1,] = c(sd(F_0.001$F),
                                                        sd(sMLH_df$sMLH_0.001),
                                                        sd(IR_df$IR_0.001),
                                                        sd(HL_df$HL_0.001))

# rename sd row
rownames(coeff_summary_0.001)[rownames(coeff_summary_0.001) == "7"] <- "sd"

# save df
write.table(coeff_summary_0.001, file = "inbreeding-coefficients/coeff_summary_0.001.tsv", sep = "\t")

# summary for live whales separately
live <- combined_data %>% filter(new_category == "Live biopsy")

coeff_summary_live <- data.frame(F = unclass(summary(live$F_0.001)),
                                 sMLH = unclass(summary(live$sMLH_0.001)),
                                 IR = unclass(summary(live$IR_0.001)),
                                 HL = unclass(summary(live$HL_0.001)))
# add sd
coeff_summary_live[nrow(coeff_summary_live) + 1,] = c(sd(live$F_0.001),
                                                      sd(live$sMLH_0.001),
                                                      sd(live$IR_0.001),
                                                      sd(live$HL_0.001))
rownames(coeff_summary_live)[rownames(coeff_summary_live) == "7"] <- "sd"

# round decimals
coeff_summary_live <- round(coeff_summary_live, 3)

# summary for dead whales separately
dead <- combined_data %>% filter(new_category != "Live biopsy")

coeff_summary_dead <- data.frame(F = unclass(summary(dead$F_0.001)),
                                 sMLH = unclass(summary(dead$sMLH_0.001)),
                                 IR = unclass(summary(dead$IR_0.001)),
                                 HL = unclass(summary(dead$HL_0.001)))
# add sd
coeff_summary_dead[nrow(coeff_summary_dead) + 1,] = c(sd(dead$F_0.001),
                                                      sd(dead$sMLH_0.001),
                                                      sd(dead$IR_0.001),
                                                      sd(dead$HL_0.001))
rownames(coeff_summary_dead)[rownames(coeff_summary_dead) == "7"] <- "sd"

# round decimals
coeff_summary_dead <- round(coeff_summary_dead, 3)

# save df
write.table(coeff_summary_live, file = "inbreeding-coefficients/coeff_summary_live.tsv", sep = "\t")
write.table(coeff_summary_dead, file = "inbreeding-coefficients/coeff_summary_dead.tsv", sep = "\t")
```

### Inbreeding coefficient histograms

I will plot inbreeding coefficient (F, sMLH, IR, HL) values in histograms for each data set:
```{r histogram of inbreeding coefficients, fig.height=16, fig.width=10}
# create plots
F_0.01_hist = ~hist(F_0.01$F, main = "0.01", col = "#677E8E", xlab = "F")
F_0.001_hist = ~hist(F_0.001$F, main = "0.001", col = "#88A2B9", xlab = "F")

sMLH_0.01_hist = ~hist(sMLH_df$sMLH_0.01, main = "0.01", col = "#677E8E", xlab = "sMLH")
sMLH_0.001_hist = ~hist(sMLH_df$sMLH_0.001, main = "0.001", col = "#88A2B9", xlab = "sMLH")

IR_0.01_hist = ~hist(IR_df$IR_0.01, main = "0.01", col = "#677E8E", xlab = "IR")
IR_0.001_hist = ~hist(IR_df$IR_0.001, main = "0.001", col = "#88A2B9", xlab = "IR")

HL_0.01_hist = ~hist(HL_df$HL_0.01, main = "0.01", col = "#677E8E", xlab = "HL")
HL_0.001_hist = ~hist(HL_df$HL_0.001, main = "0.001", col = "#88A2B9", xlab = "HL")

# save 0.01
pdf("plots/0.01/coefficient-histograms-0.01.pdf", width = 10, height = 6)
plot_grid(F_0.01_hist,
          sMLH_0.01_hist,
          IR_0.01_hist,
          HL_0.01_hist,
          nrow = 2, ncol = 2,
          labels = c("A", "B", "C", "D"), label_size = 12)
dev.off()

# save 0.001
pdf("plots/coefficient-histograms-0.001.pdf", width = 10, height = 6)
plot_grid(F_0.001_hist,
          sMLH_0.001_hist,
          IR_0.001_hist,
          HL_0.001_hist,
          nrow = 2, ncol = 2,
          labels = c("A", "B", "C", "D"), label_size = 12)
dev.off()
```

### Read depth vs. heterozygosity

Next I will plot sample mean read depth vs. inbreeding coefficients to elucidate whether heterozygosity is correlated to the sequencing read depth.  
  
First, extract mean read depth per sample data from gstacks.log.distribs files:
```
#--- make output directory ---#
mkdir ~/projects/def-frasiert/beluga-ddRAD/log-extracts/read-depth

#--- capture read depth (coverage) data from the gstacks.log.distribs files ---#
cd ~/projects/def-frasiert/beluga-ddRAD/stacks-output/with-RD-filter/gt0.01
head -n 329 gstacks.log.distribs | tail -n 160 \
  > ~/projects/def-frasiert/beluga-ddRAD/log-extracts/read-depth/0.01_read-depth

cd ~/projects/def-frasiert/beluga-ddRAD/stacks-output/with-RD-filter/gt0.001
head -n 329 gstacks.log.distribs | tail -n 160 \
  > ~/projects/def-frasiert/beluga-ddRAD/log-extracts/read-depth/0.001_read-depth
```
scp files to my computer from the cluster.  
  
Import read depth data:
```{r import read depth data}
# import mean read depth per sample data (read depth does not change based on alpha values)
# I have included the data for 0.001 separately to verify they are the same mean coverage values
read_depth = read_delim("read-depth-stats/0.01_read-depth")
read_depth_0.001 = read_delim("read-depth-stats/0.001_read-depth")
comparedf(read_depth, read_depth_0.001)

# remove replicate samples
read_depth = read_depth %>% filter(sample != "Bel00470R_FRA2297A19-2", 
                                   sample != "Bel00595R_FRA2297A22-6",
                                   sample != "Bel00711R_FRA2297A22-7",
                                   sample != "Bel00743R_FRA2297A17-8",
                                   sample != "Bel01211R_FRA2297A4-3")

# shorten sample names to just Bel numbers
read_depth$sample = substr(read_depth$sample, 1, 8)
```

Plot read depth vs. heterozygosity:
```{r plot read depth vs. heterozygosity}
# create a data frame for plot data
## first create a combined df of F values with shortened sampled names
F_df = data.frame(sample = substr(F_0.01$INDV, 1, 8), F_0.01 = F_0.01$F, F_0.001 = F_0.001$F)
## join dfs
genetic_data = read_depth %>% select(sample, n_loci, mean_cov) %>%
                              inner_join(F_df, by = join_by(sample)) %>%
                              inner_join(sMLH_df, by = join_by(sample == sample)) %>%
                              inner_join(IR_df, by = join_by(sample == sample)) %>%
                              inner_join(HL_df, by = join_by(sample == sample))

# save df to a table
write.table(genetic_data, file = "inbreeding-coefficients/inbreed_coefficients.tsv", sep = "\t", row.names = FALSE)

# plot
FvsRD_0.01 = ggplot(data = genetic_data, aes(x = mean_cov, y = F_0.01)) +
             geom_point() +
             geom_smooth(colour = "#677E8E", fill = "#677E8E") +
             theme_bw() +
             theme(text = element_text(size = 10,  family = "sans"),
                   axis.title.x = element_text(margin = margin(t = 7)),
                   axis.title.y = element_text(vjust = 2)) +
            labs(x = "Mean Read Depth", y = "F")

FvsRD_0.001 = ggplot(data = genetic_data, aes(x = mean_cov, y = F_0.001)) +
              geom_point() +
              geom_smooth(colour = "#88A2B9", fill = "#88A2B9") +
              theme_bw() +
              theme(text = element_text(size = 10,  family = "sans"),
                    axis.title.x = element_text(margin = margin(t = 7)),
                    axis.title.y = element_text(vjust = 2)) +
             labs(x = "Mean Read Depth", y = "F")

sMLHvsRD_0.01 = ggplot(data = genetic_data, aes(x = mean_cov, y = sMLH_0.01)) +
                geom_point() +
                geom_smooth(colour = "#677E8E", fill = "#677E8E") +
                theme_bw() +
                theme(text = element_text(size = 10,  family = "sans"),
                      axis.title.x = element_text(margin = margin(t = 7)),
                      axis.title.y = element_text(vjust = 2)) +
                labs(x = "Mean Read Depth", y = "sMLH")

sMLHvsRD_0.001 = ggplot(data = genetic_data, aes(x = mean_cov, y = sMLH_0.001)) +
                 geom_point() +
                 geom_smooth(colour = "#88A2B9", fill = "#88A2B9") +
                 theme_bw() +
                 theme(text = element_text(size = 10,  family = "sans"),
                       axis.title.x = element_text(margin = margin(t = 7)),
                       axis.title.y = element_text(vjust = 2)) +
                 labs(x = "Mean Read Depth", y = "sMLH")

IRvsRD_0.01 = ggplot(data = genetic_data, aes(x = mean_cov, y = IR_0.01)) +
              geom_point() +
              geom_smooth(colour = "#677E8E", fill = "#677E8E") +
              theme_bw() +
              theme(text = element_text(size = 10,  family = "sans"),
                    axis.title.x = element_text(margin = margin(t = 7)),
                    axis.title.y = element_text(vjust = 2)) +
              labs(x = "Mean Read Depth", y = "IR")

IRvsRD_0.001 = ggplot(data = genetic_data, aes(x = mean_cov, y = IR_0.001)) +
               geom_point() +
               geom_smooth(colour = "#88A2B9", fill = "#88A2B9") +
               theme_bw() +
               theme(text = element_text(size = 10,  family = "sans"),
                     axis.title.x = element_text(margin = margin(t = 7)),
                     axis.title.y = element_text(vjust = 2)) +
               labs(x = "Mean Read Depth", y = "IR")

HLvsRD_0.01 = ggplot(data = genetic_data, aes(x = mean_cov, y = HL_0.01)) +
              geom_point() +
              geom_smooth(colour = "#677E8E", fill = "#677E8E") +
              theme_bw() +
              theme(text = element_text(size = 10,  family = "sans"),
                    axis.title.x = element_text(margin = margin(t = 7)),
                    axis.title.y = element_text(vjust = 2)) +
              labs(x = "Mean Read Depth", y = "HL")

HLvsRD_0.001 = ggplot(data = genetic_data, aes(x = mean_cov, y = HL_0.001)) +
               geom_point() +
               geom_smooth(colour = "#88A2B9", fill = "#88A2B9") +
               theme_bw() +
               theme(text = element_text(size = 10,  family = "sans"),
                     axis.title.x = element_text(margin = margin(t = 7)),
                     axis.title.y = element_text(vjust = 2)) +
               labs(x = "Mean Read Depth", y = "HL")

# organize plots into a grid
pdf("plots/both_datasets/RDvsHet-plots.pdf", width = 10, height = 16)

plot_grid(FvsRD_0.01, FvsRD_0.001,
          sMLHvsRD_0.01, sMLHvsRD_0.001,
          IRvsRD_0.01, IRvsRD_0.001,
          HLvsRD_0.01, HLvsRD_0.001,
          nrow = 4, ncol = 2,
          labels = "AUTO", label_size = 12)

dev.off()

# make a grid of only 0.001 plots
pdf("plots/0.001-RDvsHet-plots.pdf", width = 10, height = 8)

plot_grid(FvsRD_0.001, sMLHvsRD_0.001,
          IRvsRD_0.001, HLvsRD_0.001,
          nrow = 2, ncol = 2,
          labels = "AUTO", label_size = 12)

dev.off()
```

### Inbreeding coefficient correlation plots

Plots of correlation between the different inbreeding coefficients:
```{r coefficient correlation plots}
IRvsF_0.01 = ggplot(data = genetic_data, aes(x = IR_0.01, y = F_0.01)) +
                    geom_point() +
                    geom_smooth(method = "lm", colour = "#677E8E", fill = "#677E8E") +
                    theme_bw() +
                    theme(text = element_text(size = 10,  family = "sans"),
                          axis.title.x = element_text(margin = margin(t = 7)),
                          axis.title.y = element_text(vjust = 2)) +
                    labs(x = "Internal relatedness (IR)", y = "F")
IRvsF_0.01

IRvsF_0.001 = ggplot(data = genetic_data, aes(x = IR_0.001, y = F_0.001)) +
                     geom_point() +
                     geom_smooth(method = "lm", colour = "#88A2B9", fill = "#88A2B9") +
                     theme_bw() +
                     theme(text = element_text(size = 10,  family = "sans"),
                           axis.title.x = element_text(margin = margin(t = 7)),
                           axis.title.y = element_text(vjust = 2)) +
                     labs(x = "Internal relatedness (IR)", y = "F")
IRvsF_0.001


IRvssMLH_0.01 = ggplot(data = genetic_data, aes(x = IR_0.01, y = sMLH_0.01)) +
                       geom_point() +
                       geom_smooth(method = "lm", colour = "#677E8E", fill = "#677E8E") +
                       theme_bw() +
                       theme(text = element_text(size = 10,  family = "sans"),
                             axis.title.x = element_text(margin = margin(t = 7)),
                             axis.title.y = element_text(vjust = 2)) +
                       labs(x = "Internal relatedness (IR)",
                            y = expression(atop("Standardized multilocus heterozygosity", "(sMLH)")))
IRvssMLH_0.01

IRvssMLH_0.001 = ggplot(data = genetic_data, aes(x = IR_0.001, y = sMLH_0.001)) +
                     geom_point() +
                     geom_smooth(method = "lm", colour = "#88A2B9", fill = "#88A2B9") +
                     theme_bw() +
                     theme(text = element_text(size = 10,  family = "sans"),
                           axis.title.x = element_text(margin = margin(t = 7)),
                           axis.title.y = element_text(vjust = 2)) +
                     labs(x = "Internal relatedness (IR)",
                          y = expression(atop("Standardized multilocus heterozygosity", "(sMLH)")))
IRvssMLH_0.001


IRvsHL_0.01 = ggplot(data = genetic_data, aes(x = IR_0.01, y = HL_0.01)) +
                       geom_point() +
                       geom_smooth(method = "lm", colour = "#677E8E", fill = "#677E8E") +
                       theme_bw() +
                       theme(text = element_text(size = 10,  family = "sans"),
                             axis.title.x = element_text(margin = margin(t = 7)),
                             axis.title.y = element_text(vjust = 2)) +
                       labs(x = "Internal relatedness (IR)",
                            y = expression(atop("Homozygosity by locus", "(HL)")))
IRvsHL_0.01

IRvsHL_0.001 = ggplot(data = genetic_data, aes(x = IR_0.001, y = HL_0.001)) +
                     geom_point() +
                     geom_smooth(method = "lm", colour = "#88A2B9", fill = "#88A2B9") +
                     theme_bw() +
                     theme(text = element_text(size = 10,  family = "sans"),
                           axis.title.x = element_text(margin = margin(t = 7)),
                           axis.title.y = element_text(vjust = 2)) +
                     labs(x = "Internal relatedness (IR)",
                          y = "Homozygosity by locus (HL)")
IRvsHL_0.001

# save plots
ggsave(plot = IRvsF_0.01, "plots/0.01/IRvsF_0.01.tiff", "tiff", width = 8.5, height = 5, units = "in")
ggsave(plot = IRvsF_0.001, "plots/IRvsF_0.001.tiff", "tiff", width = 8.5, height = 5, units = "in")
ggsave(plot = IRvssMLH_0.01, "plots/0.01/IRvssMLH_0.01.tiff", "tiff", width = 8.5, height = 5, units = "in")
ggsave(plot = IRvssMLH_0.001, "plots/IRvssMLH_0.001.tiff", "tiff", width = 8.5, height = 5, units = "in")
ggsave(plot = IRvsHL_0.01, "plots/0.01/IRvsHL_0.01.tiff", "tiff", width = 8.5, height = 5, units = "in")
ggsave(plot = IRvsHL_0.001, "plots/IRvsHL_0.001.tiff", "tiff", width = 8.5, height = 5, units = "in")

# organize plots into grid
pdf("plots/both_datasets/coefficient-correlation-plots.pdf", width = 10, height = 12)

plot_grid(IRvsF_0.01, IRvsF_0.001,
          IRvssMLH_0.01, IRvssMLH_0.001,
          IRvsHL_0.01, IRvsHL_0.001,
          nrow = 3, ncol = 2,
          labels = "AUTO", label_size = 12)

dev.off()

# make a grid of only 0.001 plots
pdf("plots/0.001-coefficient-correlation-plots.pdf", width = 15, height = 4)

plot_grid(IRvsF_0.001, IRvssMLH_0.001, IRvsHL_0.001,
          nrow = 1, ncol = 3,
          labels = "AUTO", label_size = 12)

dev.off()
```

### Sample info plot

Create a function to wrap long x-axis labels:
```{r wrap x-axis label function}
get_wrapper = function(width) {
              function(x) {
              lapply(strwrap(x, width = width, simplify = FALSE), paste, collapse="\n")
              }
}
```

Make a bar plot of sample counts and sex breakdown by mortality category:
```{r sample info bar plot, fig.height=4, fig.width=6}
sample_info_plot = ggplot(data = mortality, aes(factor(new_category))) +
                   geom_bar(stat = "count", aes(fill = sex)) +
                   theme_bw() +
                   theme(text = element_text(size = 10,  family = "sans"),
                         axis.title.x = element_text(margin = margin(t = 7)),
                         axis.title.y = element_text(vjust = 2),
                         legend.position = "bottom") +
                   labs(x = "Cause of mortality", y = "Number of individuals") +
                   scale_x_discrete(labels = get_wrapper(10)) +
                   scale_fill_manual(values = c("#decf00", "#677E8E", "#2d4030"), name = "",
                                     labels = c("Female", "Male", "M-F")) +
                   geom_text(aes(label = after_stat(count)), stat = "count", vjust = -0.25, colour = "black", size = 3)
sample_info_plot

# save plot
ggsave(plot = sample_info_plot, "plots/sample_info_plot.tiff", "tiff", width = 6, height = 4, units = "in")
```

### Inbreeding coefficient dotplots divided by sex

I will create another mortality category that splits up infection type.
```{r separate infection categories - Figure 1}
# create updated mortality class that divides infectious disease by type: verminous infection, protozoan infection, bacterial infection
## combining undetermined & viral infections into other infection due to sample size
combined_data = combined_data %>% mutate(split_inf_category = case_when(
                new_category == "Dystocia" ~ "Dystocia",
                new_category == "Anthropogenic causes" ~ "Anthropogenic causes",
                new_category == "Neoplasia" ~ "Neoplasia",
                new_category == "Degenerative changes" ~ "Degenerative changes",
                new_category == "Undetermined" ~ "Undetermined",
                new_category == "Neonatal mortality" ~ "Neonatal mortality",
                new_category == "Live biopsy" ~ "Live biopsy",
                category == "Infectious - bacterial" ~ "Bacterial infection",
                category == "Infectious - bactérien" ~ "Bacterial infection",
                category == "Infectious - undetermined" ~ "Other infection",
                category == "Infectious - viral" ~ "Other infection",
                category == "Protozoal pneumonia" ~ "Protozoan infection",
                category == "Toxoplasmosis" ~ "Protozoan infection",
                category == "Verminous bronchopneumonia" ~ "Verminous infection",
                category == "Verminous gastro-enteritis / peritonitis" ~ "Verminous infection"))

# reorder the new category factor to put live biopsy last & to make other infectious the last in that group
combined_data$split_inf_category = factor(combined_data$split_inf_category, levels = c("Anthropogenic causes", "Bacterial infection", "Degenerative changes", "Dystocia", "Neonatal mortality",  "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection","Undetermined", "Live biopsy"))

# count n per category
combined_data %>% dplyr::count(split_inf_category)

# save this df
write.table(combined_data, file = "combined-data.tsv", sep = "\t", row.names = FALSE)

# make another sample info plot
sample_info_plot2 = ggplot(data = combined_data %>% filter((split_inf_category != "Live biopsy")),
                           aes(factor(split_inf_category))) +
                    geom_bar(stat = "count", aes(fill = sex)) +
                    theme_bw() +
                    theme(text = element_text(size = 8.125,  family = "sans"),
                          axis.title = element_text(size = 8.125),
                          axis.title.x = element_text(margin = margin(t = -2, b = -6.5)),
                          axis.title.y = element_text(vjust = 2),
                          axis.text.x = element_text(angle = 90, vjust = 0.5,
                                                     hjust = 1),
                          legend.position = "bottom",
                          legend.key.size = unit(2.5, 'mm'),
                          legend.margin = margin(b = 0),
                          legend.text = element_text(margin = margin(l = 1.25, r = 1.25))) +
                    labs(x = "Cause of mortality", y = "Number of individuals") +
                    scale_x_discrete(labels = get_wrapper(10)) +
                    ylim(0, 33.5) +
                    scale_fill_manual(values = c("#decf00", "#677E8E", "#2d4030"), name = "",
                                      labels = c("Female", "Male", "M-F")) +
                    geom_text(aes(label = after_stat(count)), stat = "count", vjust = -0.25, colour = "black", size = 1.95)
sample_info_plot2

# save plot
ggsave(plot = sample_info_plot2, "plots/sample_info_plot2.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = sample_info_plot2, "../figures/Figure_1.pdf", "pdf", width = 80, height = 65, units = "mm", dpi = 1000)
```

I will plot inbreeding coefficient (F, sMLH, IR, HL) values in dotplots for the 0.001 dataset:
```{r dotplots of inbreeding coefficients for 0.001 - Figure 2, fig.height=4, fig.width=6}
F_dotplot_0.001 = ggplot(data = combined_data, aes(x = F_0.001, fill = sex)) +
                  geom_dotplot(method = "histodot",
                               stackgroups = TRUE, binpositions = "all", binwidth = 0.0122705) +
                  theme_bw() +
                  theme(text = element_text(size = 10,  family = "sans"),
                        axis.title.x = element_text(margin = margin(t = 3)),
                        axis.title.y = element_blank(),
                        axis.text.y = element_blank(),
                        axis.ticks.y = element_blank(),
                        legend.position = "none",
                        plot.margin = margin(b = 3, t = 5, l = 3, r = 2)) +
                  scale_fill_manual(values = c("#decf00", "#677E8E", "#2d4030")) +
                  labs(x = "F")
F_dotplot_0.001

sMLH_dotplot_0.001 = ggplot(data = combined_data, aes(x = sMLH_0.001, fill = sex)) +
                    geom_dotplot(method = "histodot",
                                 stackgroups = TRUE, binpositions = "all", binwidth = 0.0108) +
                    theme_bw() +
                    theme(text = element_text(size = 10,  family = "sans"),
                          axis.title.x = element_text(margin = margin(t = 3)),
                          axis.title.y = element_blank(),
                          axis.text.y = element_blank(),
                          axis.ticks.y = element_blank(),
                          legend.position = "none",
                          plot.margin = margin(b = 3, t = 5,  l = 2, r = 5)) +
                    scale_fill_manual(values = c("#decf00", "#677E8E", "#2d4030")) +
                    scale_x_continuous(breaks = c(0.9, 1.0, 1.1, 1.2)) +
                    labs(x = "sMLH")
sMLH_dotplot_0.001

IR_dotplot_0.001 = ggplot(data = combined_data, aes(x = IR_0.001, fill = sex)) +
                   geom_dotplot(method = "histodot",
                                stackgroups = TRUE, binpositions = "all", binwidth = 0.010012) +
                   theme_bw() +
                   theme(text = element_text(size = 10,  family = "sans"),
                         axis.title.x = element_text(margin = margin(t = 3)),
                         axis.title.y = element_blank(),
                         axis.text.y = element_blank(),
                         axis.ticks.y = element_blank(),
                         legend.position = "none",
                         plot.margin = margin(b = 4, l = 3, r = 2, t = 5)) +
                   scale_fill_manual(values = c("#decf00", "#677E8E", "#2d4030")) +
                   labs(x = "IR")
IR_dotplot_0.001

HL_dotplot_0.001 = ggplot(data = combined_data, aes(x = HL_0.001, fill = sex)) +
                   geom_dotplot(method = "histodot",
                                stackgroups = TRUE, binpositions = "all", binwidth = 0.004) +
                   theme_bw() +
                   theme(text = element_text(size = 10,  family = "sans"),
                         axis.title.x = element_text(margin = margin(t = 3)),
                         axis.title.y = element_blank(),
                         axis.text.y = element_blank(),
                         axis.ticks.y = element_blank(),
                         legend.position = "none",
                         plot.margin = margin(b = 4, l = 2, r = 5, t = 5)) +
                   scale_fill_manual(values = c("#decf00", "#677E8E", "#2d4030")) +
                   labs(x = "HL")
HL_dotplot_0.001

# make a grid of the plots
histo_grid <- plot_grid(F_dotplot_0.001, sMLH_dotplot_0.001,
              IR_dotplot_0.001, HL_dotplot_0.001,
              nrow = 2, ncol = 2,
              labels = c("(a)","(b)","(c)","(d)"), label_size = 12, hjust = -0.6, vjust = 2)
histo_grid

ggsave(plot = histo_grid, "plots/0.001-coefficient-dotplots.pdf", width = 12, height = 8)
ggsave(plot = histo_grid, "../figures/Figure_2.pdf", "pdf", width = 180, height = 120, units = "mm", dpi = 1000)
```

### Cause of death vs. inbreeding coefficient

Box plot of IR vs. cause of mortality:
```{r IR vs. COD box plot, fig.height=4, fig.width=6}
# create a data frame to label the tops of bars
n_label = combined_data %>% dplyr::count(new_category)
# add "n = " to the start of the labels
n_label$n = paste("n=", n_label$n, sep = "")

# create a colour palette to highlight live biopsy samples in a different colour
biopsy_colour = c("black", "black", "black", "black", "black", "black", "black", "#88A2B9")
# create a face palette to highlight live biopsy labels in bold
biopsy_face = c("plain", "plain", "plain", "plain", "plain", "plain", "plain", "bold")

# plot
IR_boxplot_0.001_sex = ggplot(data = combined_data, aes(x = factor(new_category), y = IR_0.001)) +
                       geom_boxplot(outlier.shape = NA) +
                       geom_boxplot(data = combined_data[combined_data$new_category == "Live biopsy", ],
                                    outlier.shape = NA, aes(x = new_category, y = IR_0.001), colour = "#88A2B9") +
                       geom_jitter(alpha = 0.25,
                                   aes(colour = new_category, shape = sex, stroke = 1.5),
                                   width = 0.25) +
                       scale_color_manual(values = biopsy_colour, guide = "none") +
                       scale_shape_discrete(labels = c("Female","Male", "M-F")) +
                       theme_bw() +
                       theme(text = element_text(size = 12,  family = "sans"),
                             legend.position = "right",
                             axis.title.x = element_text(margin = margin(t = 7)),
                             axis.title.y = element_text(vjust = 2),
                             axis.text.x = element_text(colour = biopsy_colour, face = biopsy_face,
                                                        )) +
                       labs(x = "Sample category", y = "Internal relatedness (IR)", shape = NULL, legend = NULL) +
                       scale_x_discrete(labels = get_wrapper(10)) +
                       annotate("text", x = n_label$new_category, y = 0.06,
                                label = n_label$n,
                                colour = biopsy_colour,
                                fontface = c(1, 1, 1, 1, 1, 1, 1, 2), family = "sans") +
                       theme(legend.box.spacing = unit(0, "pt"),
                             legend.spacing = unit(0, "mm"),
                             legend.spacing.x = unit(0, 'mm'))
IR_boxplot_0.001_sex
```

Box plot of IR vs. cause of mortality that includes separate infection causes and sex:
```{r IR boxlot with sex AND separate infection causes - Figure 3}
# create a data frame to label the tops of bars
n_label_inf = combined_data %>% dplyr::count(split_inf_category)

# add "n = " to the start of the labels
n_label_inf$n = paste("n=", n_label_inf$n, sep = "")

# create a colour palette to highlight live biopsy samples in a different colour
biopsy_colour2 = c("black", "black", "black", "black", "black", "black", "black", "black", "black", "black","#88A2B9")
# create a face palette to highlight live biopsy labels in bold
#biopsy_face2 = c("plain", "plain", "plain", "plain", "plain", "plain", "plain", "plain", "plain", "plain", "bold")

# same plot, but now with separate infectious categories
IR_boxplot_0.001_SplitInf = ggplot(data = combined_data, aes(x = factor(split_inf_category), y = IR_0.001)) +
                            geom_boxplot(outlier.shape = NA, size = 0.45) +
                            geom_boxplot(data = combined_data[combined_data$split_inf_category == "Live biopsy", ],
                                         outlier.shape = NA, aes(x = split_inf_category,
                                                                 y = IR_0.001),
                                         colour = "#88A2B9", size = 0.45) +
                            geom_jitter(data = combined_data[combined_data$split_inf_category != "Live biopsy", ],
                                        alpha = 0.26,
                                        aes(colour = split_inf_category, shape = sex,
                                            stroke = 0.24),
                                        width = 0.2) +
                            geom_jitter(data = combined_data[combined_data$split_inf_category == "Live biopsy", ],
                                        alpha = 0.65,
                                        aes(colour = split_inf_category, shape = sex,
                                            stroke = 0.24),
                                        width = 0.2) +
                            scale_color_manual(values = biopsy_colour2, guide = "none") +
                            scale_shape_discrete(labels = c("Female","Male", "M-F")) +
                            theme_bw() +
                            theme(text = element_text(size = 8.125,  family = "sans"),
                                  axis.title = element_text(size = 8.2),
                                  legend.position = "bottom",
                                  legend.key.size = unit(2, 'mm'),
                                  legend.text = element_text(margin = margin(l = 2, r = 2)),
                                  legend.margin = margin(b = 0),
                                  axis.title.x = element_text(margin = margin(t = -2, b = -6.5)),
                                  axis.title.y = element_text(vjust = 2),
                                  axis.text.x = element_text(colour = biopsy_colour2, #face = biopsy_face2,
                                                angle = 90, vjust = 0.5, hjust = 1)) +
                            labs(x = "Sample category", y = "Internal relatedness (IR)",
                                shape = NULL, legend = NULL) +
                            scale_x_discrete(labels = get_wrapper(11)) #+
                            #annotate("text", x = n_label_inf$split_inf_category, y = 0.07,
                                     #label = n_label_inf$n,
                                     #colour = biopsy_colour2,
                                     #fontface = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2),
                                     #family = "sans",
                                     #size = 1.9)

IR_boxplot_0.001_SplitInf

# save plot
ggsave(plot = IR_boxplot_0.001_SplitInf, "plots/SplitInf_IR_boxplot_0.001.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = IR_boxplot_0.001_SplitInf, "../figures/Figure_3.pdf", "pdf", width = 80, height = 70, units = "mm", dpi = 1000)
```

Box plot of sMLH vs. cause of mortality:
```{r sMLH vs. COD box plot, fig.height=5, fig.width=8.5}
sMLH_boxplot_0.001 = ggplot(data = combined_data, aes(x = factor(new_category), y = sMLH_0.001)) +
                     geom_boxplot(outlier.shape = NA) +
                     geom_boxplot(data = combined_data[combined_data$new_category == "Live biopsy", ],
                                  outlier.shape = NA, aes(x = new_category, y = sMLH_0.001), colour = "#88A2B9") +
                     geom_jitter(alpha = 0.25, aes(colour = new_category), width = 0.25) +
                     scale_color_manual(values = biopsy_colour) +
                     theme_bw() +
                     theme(text = element_text(size = 12,  family = "sans"),
                           legend.position = "none",
                           axis.title.x = element_text(margin = margin(t = 7)),
                           axis.title.y = element_text(vjust = 2),
                           axis.text.x = element_text(colour = biopsy_colour, face = biopsy_face,)) +
                     labs(x = "Sample category",
                          y = expression(atop("Standardized multilocus heterozygosity", "(sMLH)"))) +
                     scale_x_discrete(labels = get_wrapper(10)) +
                     annotate("text", x = n_label$new_category, y = 1.29,
                              label = n_label$n,
                              colour = biopsy_colour,
                              fontface = c(1, 1, 1, 1, 1, 1, 1, 2), family = "sans")
sMLH_boxplot_0.001
ggsave(plot = sMLH_boxplot_0.001, "plots/sMLH_boxplot_0.001.tiff", "tiff", width = 6, height = 4, units = "in")
```

Box plot of HL vs. cause of mortality:
```{r HL vs. COD box plot, fig.height=5, fig.width=8.5}
HL_boxplot_0.001 = ggplot(data = combined_data, aes(x = factor(new_category), y = HL_0.001)) +
                   geom_boxplot(outlier.shape = NA) +
                   geom_boxplot(data = combined_data[combined_data$new_category == "Live biopsy", ],
                                outlier.shape = NA, aes(x = new_category, y = HL_0.001), colour = "#88A2B9") +
                   geom_jitter(alpha = 0.25, aes(colour = new_category), width = 0.25) +
                   scale_color_manual(values = biopsy_colour) +
                   theme_bw() +
                   theme(text = element_text(size = 12,  family = "sans"),
                         legend.position = "none",
                         axis.title.x = element_text(margin = margin(t = 7)),
                         axis.title.y = element_text(vjust = 2),
                         axis.text.x = element_text(colour = biopsy_colour, face = biopsy_face,)) +
                   labs(x = "Sample category", y = "Homozygosity by locus (HL)") +
                   scale_x_discrete(labels = get_wrapper(10)) +
                   annotate("text", x = n_label$new_category, y = 0.677,
                            label = n_label$n,
                            colour = biopsy_colour,
                            fontface = c(1, 1, 1, 1, 1, 1, 1, 2), family = "sans")
HL_boxplot_0.001

# save plot
ggsave(plot = HL_boxplot_0.001, "plots/extra_plots/HL_boxplot_0.001.tiff", "tiff", width = 6, height = 4, units = "in")
```

### Age vs. inbreeding coefficients

```{r age vs. inbreeding plots - Figure 5, fig.height=4, fig.width=6}
# plot age vs. IR using the 0.001 dataset
age_vs_IR_0.001 = ggplot(data = (combined_data %>% filter(new_category != "Live biopsy")),
                         aes(x = estimated_age, y = IR_0.001)) +
                  geom_point() +
                  geom_smooth(method = "lm", colour = "#88A2B9", fill = "#88A2B9") +
                  theme_bw() +
                  theme(text = element_text(size = 12,  family = "sans"),
                        axis.title.x = element_text(margin = margin(t = 7)),
                        axis.title.y = element_text(vjust = 2))+
                  labs(x = "Age (years)", y = "IR")
age_vs_IR_0.001

# plot age vs. sMLH using the 0.001 dataset
age_vs_sMLH_0.001 = ggplot(data = (combined_data %>% filter(new_category != "Live biopsy")),
                         aes(x = estimated_age, y = sMLH_0.001)) +
                  geom_point() +
                  geom_smooth(method = "lm", colour = "#88A2B9", fill = "#88A2B9") +
                  theme_bw() +
                  theme(text = element_text(size = 12,  family = "sans"),
                        axis.title.x = element_text(margin = margin(t = 7)),
                        axis.title.y = element_text(vjust = 2))+
                  labs(x = "Age (years)", y = "sMLH")
age_vs_sMLH_0.001

# plot age vs. HL using the 0.001 dataset
age_vs_HL_0.001 = ggplot(data = (combined_data %>% filter(new_category != "Live biopsy")),
                         aes(x = estimated_age, y = HL_0.001)) +
                  geom_point() +
                  geom_smooth(method = "lm", colour = "#88A2B9", fill = "#88A2B9") +
                  theme_bw() +
                  theme(text = element_text(size = 12,  family = "sans"),
                        axis.title.x = element_text(margin = margin(t = 7)),
                        axis.title.y = element_text(vjust = 2))+
                  labs(x = "Age (years)", y = "HL")
age_vs_HL_0.001

# plot age vs. F using the 0.001 dataset
age_vs_F_0.001 = ggplot(data = (combined_data %>% filter(new_category != "Live biopsy")),
                        aes(x = estimated_age, y = F_0.001)) +
                 geom_point() +
                 geom_smooth(method = "lm", colour = "#88A2B9", fill = "#88A2B9") +
                 theme_bw() +
                 theme(text = element_text(size = 12,  family = "sans"),
                       axis.title.x = element_text(margin = margin(t = 7)),
                       axis.title.y = element_text(vjust = 2))+
                 labs(x = "Age (years)", y = "F")
age_vs_F_0.001

# create a stacked facet plot
## pivot longer
age_inb_long <- combined_data %>%
                filter(new_category != "Live biopsy") %>%
                select(sample, estimated_age, F_0.001, sMLH_0.001, IR_0.001, HL_0.001) %>%
                pivot_longer(cols = c(F_0.001, sMLH_0.001, IR_0.001, HL_0.001),
                             names_to = "inb_measure",
                             values_to = "value")
## clean up names
age_inb_long$inb_measure <- factor(age_inb_long$inb_measure,
                                   levels = c("F_0.001", "sMLH_0.001", "IR_0.001", "HL_0.001"),
                                   labels = c("F", "sMLH", "IR", "HL"))
  
age_vs_inbreeding <- ggplot(age_inb_long, aes(x = estimated_age, y = value)) +
                     geom_point(size = 0.6) +
                     facet_grid(inb_measure ~ ., scales = "free_y", switch = "y") +
                     geom_smooth(method = "lm", se = TRUE, colour = "#88A2B9", fill = "#88A2B9", size = 0.75) +
                     theme_bw() +
                     theme(text = element_text(size = 10,  family = "sans"),
                           strip.placement = "outside", strip.background = element_blank()) +
                     labs(x = "Age (years)", y = NULL) +
                     scale_x_continuous(breaks = seq(0, 65, 10),
                                        name = "Age (years)",
                                        sec.axis = dup_axis(name = ""))

age_vs_inbreeding

# save plots
ggsave(plot = age_vs_IR_0.001, "plots/age_vs_IR_0.001.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = age_vs_sMLH_0.001, "plots/age_vs_sMLH_0.001.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = age_vs_HL_0.001, "plots/age_vs_HL_0.001.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = age_vs_F_0.001, "plots/age_vs_F_0.001.tiff", "tiff", width = 6, height = 4, units = "in")

ggsave(plot = age_vs_inbreeding, "plots/age_vs_inbreeding.tiff", "tiff", width = 6, height = 10, units = "in")
ggsave(plot = age_vs_inbreeding, "../figures/Figure_5.pdf", "pdf", width = 80, height = 140, units = "mm")

# correlation tests
cor.test_IR = cor.test(combined_data$IR_0.001, combined_data$estimated_age)
cor.test_IR$estimate
cor.test_IR$p.value

cor.test_sMLH = cor.test(combined_data$sMLH_0.001, combined_data$estimated_age)
cor.test_sMLH$estimate
cor.test_sMLH$p.value
```

### Plot of COD vs. year found dead

Box plot of year found vs. cause of mortality that includes separate infection causes and sex:
```{r year boxlot with sex AND separate infection causes}
# filter out live biopsies to have just dead whales
dead_combined_data = combined_data %>% filter(split_inf_category != "Live biopsy")

# create a data frame to label the tops of bars
n_label_dead = dead_combined_data %>% dplyr::count(split_inf_category)

# add "n = " to the start of the labels
n_label_dead$n = paste("n=", n_label_dead$n, sep = "")

# plot
year_boxplot_SplitInf = ggplot(data = dead_combined_data, aes(x = factor(split_inf_category),
                                                              y = IR_0.001)) +
                               geom_boxplot(outlier.shape = NA) +
                               geom_jitter(alpha = 0.25,
                                           aes(shape = sex, stroke = 1.5),
                                           width = 0.25) +
                               scale_shape_discrete(labels = c("Female","Male", "M-F")) +
                               theme_bw() +
                               theme(text = element_text(size = 12,  family = "sans"),
                                     legend.position = "right",
                                     axis.title.x = element_text(margin = margin(t = 7)),
                                     axis.title.y = element_text(vjust = 2),
                                     axis.text.x = element_text(angle = 90, vjust = 0.5,
                                                                hjust = 1)) +
                               labs(x = "Cause of mortality", y = "Year found dead",
                                    shape = "", legend = NULL) +
                               scale_x_discrete(labels = get_wrapper(11)) +
                               annotate("text", x = n_label_dead$split_inf_category, y = 0.06,
                                        label = n_label_dead$n,
                                        family = "sans") +
                               theme(legend.box.spacing = unit(0, "pt"),
                                     legend.margin = unit(0, "mm"),
                                     legend.spacing.x = unit(0, 'mm'))
year_boxplot_SplitInf

# save plot
ggsave(plot = year_boxplot_SplitInf, "plots/year_boxplot_SplitInf.tiff", "tiff", width = 6, height = 4, units = "in")
```

Box plot of age vs. cause of mortality that includes separate infection causes and sex:
```{r age boxlot with sex AND separate infection causes}
age_boxplot_SplitInf = ggplot(data = dead_combined_data, aes(x = factor(split_inf_category), y = IR_0.001)) +
                              geom_boxplot(outlier.shape = NA) +
                              geom_jitter(alpha = 0.25,
                                          aes(shape = sex, stroke = 1.5),
                                          width = 0.25) +
                              scale_shape_discrete(labels = c("Female","Male", "M-F")) +
                              theme_bw() +
                              theme(text = element_text(size = 12,  family = "sans"),
                                    legend.position = "right",
                                    axis.title.x = element_text(margin = margin(t = 7)),
                                    axis.title.y = element_text(vjust = 2),
                                    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
                              labs(x = "Cause of mortality", y = "Year found dead", shape = "", legend = NULL) +
                              scale_x_discrete(labels = get_wrapper(11)) +
                              annotate("text", x = n_label_dead$split_inf_category, y = 0.06,
                                       label = n_label_dead$n,
                                       family = "sans") +
                              theme(legend.box.spacing = unit(0, "pt"),
                                    legend.margin = unit(0, "mm"),
                                    legend.spacing.x = unit(0, 'mm'))

year_boxplot_SplitInf

# save plot
ggsave(plot = year_boxplot_SplitInf, "plots/year_boxplot_SplitInf.tiff", "tiff", width = 6, height = 4, units = "in")
```

## 3. Bayesian models

### Objectives

I want to use Bayesian methods answer these 5 questions:    
1. What is the effect of inbreeding (internal relatedness) on cause of mortality?  
2. What is the effect of year of death on cause of mortality?  
3. What is the effect of age on cause of mortality?  
4. Does year of death influence the effect of inbreeding on cause of mortality?
5. Does age influence the effect of inbreeding on cause of mortality?

### Organize the data for Bayesian models

I want to pull out the year from the "date found" data column, which does not have consistent formatting.
```{r extract year found dead}
# create a new column with just year found
combined_data$year_found = str_extract(combined_data$date_found, "\\b\\d{4}\\b")

# check the parsing
combined_data %>% select(c(date_found, year_found))
```
  
*I am excluding live biopsies at this time, to focus just on the effects of inbreeding on cause of DEATH.*  I will also remove the "degenerative changes" mortality category because it contains only one individual.
```{r data filter}
model_data = combined_data %>% filter(split_inf_category != "Degenerative changes") %>%
                               filter(split_inf_category != "Live biopsy")
```

I will make the nominal predicted variable (cause of death) a factor.  
I will centre the metric predictors (age, year found, and inbreeding coefficient) around zero. 
```{r data preparation}
# nominal predicted variable is cause of mortality
# make it a factor and order to put undetermined last
model_data$mortality = factor(model_data$split_inf_category,
                              levels = c("Anthropogenic causes",
                                         "Bacterial infection",
                                         "Dystocia",
                                         "Neonatal mortality",
                                         "Neoplasia",
                                         "Other infection",
                                         "Protozoan infection",
                                         "Verminous infection",
                                         "Undetermined"))

#--- prepare and standardize metric predictor variables ---#

# age
model_data$cAge = model_data$estimated_age - mean(model_data$estimated_age)

# year found
model_data$cYear = as.numeric(model_data$year_found) - mean(as.numeric(model_data$year_found))

# inbreeding coefficients (I will standardize all coeffcients, but I am focusing on running models with IR)
model_data$cIR = model_data$IR_0.001 - mean(model_data$IR_0.001)
model_data$csMLH = model_data$sMLH_0.001 - mean(model_data$sMLH_0.001)
model_data$cF = model_data$F_0.001 - mean(model_data$F_0.001)
model_data$cHL = model_data$HL_0.001 - mean(model_data$HL_0.001)
```

### Model 1: Effect of internal relatedness (IR) on cause of mortality

```{r model1 - IR (Figure 4)}
# model
model1 = brm(formula = mortality ~ 1 + cIR, 
             data = model_data,
             family = categorical(link = logit))

# evaluate model performance
summary(model1)
mcmc_trace(model1)

# posterior predictive check
pp_check(model1, ndraws = 100, type = "bars")

# evaluate results
mcmc_intervals(model1, regex_pars = "_cIR")
mcmc_areas(model1, regex_pars = "_cIR")
mcmc_areas_ridges(model1, regex_pars = "_cIR")

posterior_summary(model1, probs = c(0.055, 0.945))

# extract the data
model1_post = as_draws_df(model1, variable = "_cIR", regex = TRUE)
posterior::summarize_draws(model1_post)

model1_slope_post = model1_post[, 1:8]
colnames(model1_slope_post) = c("Bacterial infection", "Dystocia",
                                "Neonatal mortality", "Neoplasia",
                                "Other infection", "Protozoan infection",
                                "Verminous infection", "Undetermined")

# plot slope posteriors
model1_slope_tidy = pivot_longer(model1_slope_post,
                                 cols = everything(),
                                 names_to = "Mortality")

model1_IR_ridges = ggplot(model1_slope_tidy,
                          aes(x = value, y = Mortality,
                              fill = factor(stat(quantile)))) +
                   theme_bw() +
                   theme(text = element_text(size = 17,  family = "sans")) +
                   stat_density_ridges(geom = "density_ridges_gradient",
                                       calc_ecdf = TRUE,
                                       quantiles = c(0.025, 0.975)) +
                   scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                     guide = "none") +
                   geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                   labs(#title = "Model 1: mortality ~ cIR",
                        x = "Effect of IR on mortality (log odds)",
                        y = "Cause of mortality") +
                   scale_y_discrete(expand = expansion(add = c(0.5, 2)))
model1_IR_ridges

# save plot
ggsave(plot = model1_IR_ridges, "plots/bayesian_models/model1_IR.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model1_IR_ridges, "plots/bayesian_models/Figure_4_180mm.pdf", "pdf", width = 180, height = 120, units = "mm")

# resize plot to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/Figure_4_180mm.pdf"
outfile <- "../figures/Figure_4.pdf"

target_width_mm  <- 80
target_height_mm <- 54

mm_to_pt <- 2.83465
w_pt <- target_width_mm * mm_to_pt
h_pt <- target_height_mm * mm_to_pt

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

### Model 2: Effect of year on cause of mortality

To get at the potential effects of changing environment and differential exposure to stressors over time on cause of mortality, I will look at the effect of year found dead on the probability of dying from different causes.

```{r model2 - year (Figure 8)}
# model
model2 =  brm(formula = mortality ~ 1 + cYear, 
              data = model_data, 
              family = categorical(link = logit))

# evaluate model performance
summary(model2)
mcmc_trace(model2)

# posterior predictive check
pp_check(model2, ndraws = 100, type = "bars")

# evaluate results
mcmc_intervals(model2, regex_pars = "_cYear")
mcmc_areas(model2, regex_pars = "_cYear")
mcmc_areas_ridges(model2, regex_pars = "_cYear")

posterior_summary(model2, probs = c(0.055, 0.945))

# extract the data
model2_post = as_draws_df(model2, variable = "_cYear", regex = TRUE)
posterior::summarize_draws(model2_post)

model2_slope_post = model2_post[, 1:8]
colnames(model2_slope_post) = c("Bacterial infection", "Dystocia",
                                "Neonatal mortality", "Neoplasia",
                                "Other infection", "Protozoan infection",
                                "Verminous infection", "Undetermined")

# plot slope posteriors
model2_slope_tidy = pivot_longer(model2_slope_post,
                                 cols = everything(),
                                 names_to = "Mortality")

model2_year_ridges = ggplot(model2_slope_tidy,
                            aes(x = value, y = Mortality,
                                fill = factor(stat(quantile)))) +
                     theme_bw() +
                     theme(text = element_text(size = 17,  family = "sans")) +
                     stat_density_ridges(geom = "density_ridges_gradient",
                                         calc_ecdf = TRUE,
                                         quantiles = c(0.025, 0.975)) +
                     scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                       guide = "none") +
                     geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                     labs(#title = "Model 2: mortality ~ cYear",
                          x = "Effect of year on mortality (log odds)",
                          y = "Cause of mortality") +
                     scale_y_discrete(expand = expansion(add = c(0.5, 2)))

model2_year_ridges

# save plot
ggsave(plot = model2_year_ridges, "plots/bayesian_models/model2_year.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model2_year_ridges, "plots/bayesian_models/Figure_8_180mm.pdf", "pdf", width = 180, height = 120, units = "mm")

# resize plot to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/Figure_8_180mm.pdf"
outfile <- "../figures/Figure_8.pdf"

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

### Model 3: Effect of age on cause of mortality

I expect younger individuals to be more susceptible to different causes of death from older individuals.  
  
I will run this model two ways: first with the whole dataset, and then after filtering to remove the neonatal mortality category because this category contains only individuals 0 y/o by definition and may mask the effects of age on the other causes of death. 

#### Age with all data

```{r model3 - age w/ all data}
model3 =  brm(formula = mortality ~ 1 + cAge, 
              data = model_data, 
              family = categorical(link = logit))

# evaluate model performance
summary(model3)
mcmc_trace(model3)

# posterior predictive check
pp_check(model3, ndraws = 100, type = "bars")

# evaluate results
mcmc_intervals(model3, regex_pars = "_cAge")
mcmc_areas(model3, regex_pars = "_cAge")
mcmc_areas_ridges(model3, regex_pars = "_cAge")

posterior_summary(model3, probs = c(0.055, 0.945))

# extract the data
model3_post = as_draws_df(model3, variable = "_cAge", regex = TRUE)
posterior::summarize_draws(model3_post)

model3_slope_post = model3_post[, 1:8]
colnames(model3_slope_post) = c("Bacterial infection", "Dystocia",
                                "Neonatal mortality", "Neoplasia",
                                "Other infection", "Protozoan infection",
                                "Verminous infection", "Undetermined")

# plot slope posteriors
model3_slope_tidy = pivot_longer(model3_slope_post,
                                 cols = everything(),
                                 names_to = "Mortality")

model3_age_ridges = ggplot(model3_slope_tidy,
                           aes(x = value, y = Mortality,
                               fill = factor(stat(quantile)))) +
                    theme_bw() +
                    theme(text = element_text(size = 12,  family = "sans")) +
                    stat_density_ridges(geom = "density_ridges_gradient",
                                        calc_ecdf = TRUE,
                                        quantiles = c(0.025, 0.975)) +
                    scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                      guide = "none") +
                    geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") +
                    labs(title = "Model 3: mortality ~ cAge (all data)",
                         x = "log odds (age) relative to 'anthropogenic causes'",
                         y = "Cause of mortality")
model3_age_ridges

# save plot
ggsave(plot = model3_age_ridges, "plots/bayesian_models/model3_age_alldata.tiff", "tiff", width = 6, height = 4, units = "in")
```

#### Age without neonatal mortality

```{r model3.1 - age w/o neonatal mortality (Figure 6)}
# model
model3.1 =  brm(formula = mortality ~ 1 + cAge, 
                data = (model_data %>% filter(mortality != "Neonatal mortality")), 
                family = categorical(link = logit))

# evaluate model performance
summary(model3.1)
mcmc_trace(model3.1)

# posterior predictive check
pp_check(model3.1, ndraws = 100, type = "bars")

# evaluate results
mcmc_intervals(model3.1, regex_pars = "_cAge")
mcmc_areas(model3.1, regex_pars = "_cAge")
mcmc_areas_ridges(model3.1, regex_pars = "_cAge")

posterior_summary(model3.1, probs = c(0.055, 0.945))

# extract the data
model3.1_post = as_draws_df(model3.1, variable = "_cAge", regex = TRUE)
posterior::summarize_draws(model3.1_post)

model3.1_slope_post = model3.1_post[, 1:7]
colnames(model3.1_slope_post) = c("Bacterial infection", "Dystocia",
                                  "Neoplasia", "Other infection",
                                  "Protozoan infection", "Verminous infection",
                                  "Undetermined")

# plot slope posteriors
model3.1_slope_tidy = pivot_longer(model3.1_slope_post,
                                   cols = everything(),
                                   names_to = "Mortality")

model3.1_age_ridges = ggplot(model3.1_slope_tidy,
                             aes(x = value, y = Mortality,
                                 fill = factor(stat(quantile)))) +
                      theme_bw() +
                      theme(text = element_text(size = 17,  family = "sans")) +
                      stat_density_ridges(geom = "density_ridges_gradient",
                                          calc_ecdf = TRUE,
                                          quantiles = c(0.025, 0.975)) +
                      scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                        guide = "none") +
                      geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") +           
                      labs(#title = "Model 3.1: mortality ~ cAge (w/o neonatal mortality)",
                           x = "Effect of age on mortality (log odds)",
                           y = "Cause of mortality") +
                      scale_y_discrete(expand = expansion(add = c(0.5, 2)))
model3.1_age_ridges

# save plot
ggsave(plot = model3.1_age_ridges, "plots/bayesian_models/model3.1_age_noneonatal.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model3.1_age_ridges, "plots/bayesian_models/Figure_6_180mm.pdf", "pdf", width = 180, height = 120, units = "mm")

# resize plot to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/Figure_6_180mm.pdf"
outfile <- "../figures/Figure_6.pdf"

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

### Model 4: Effect of the INTERACTION between IR and year found dead on cause of mortality

Because inbreeding depression may depend on environmental challenges (i.e., genetic differences in fitness are conditional and only revealed under adverse environmental conditions), I wonder if the year found dead (and relevant environmental exposures) has an impact on the effect of inbreeding on cause of mortality. To test this, I will implement a model with an interaction between these two variables.

```{r model 4 - IR x year (Figure 9)}
# model
model4 = brm(formula = mortality ~ 1 + cIR + cYear + cIR:cYear, 
             data = model_data, 
             family = categorical(link = logit))


stancode(model4)

# evaluate model performance 
summary(model4)

# posterior predictive check
pp_check(model4, ndraws = 100, type = "bars")

#--- evaluate results ---#
posterior_summary(model4, probs = c(0.055, 0.945))

# IR
mcmc_intervals(model4, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

mcmc_areas(model4, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

mcmc_areas_ridges(model4, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

# year
mcmc_intervals(model4, regex_pars = "_cYear")
mcmc_areas(model4, regex_pars = "_cYear")
mcmc_areas_ridges(model4, regex_pars = "_cYear")

# interaction, IR x year
mcmc_intervals(model4, regex_pars = "_cIR:cYear")
mcmc_areas(model4, regex_pars = "_cIR:cYear")
mcmc_areas_ridges(model4, regex_pars = "_cIR:cYear")

# extract the data
model4_IR_post = as_draws_df(model4, variable = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

model4_year_post = as_draws_df(model4, variable = "_cYear", regex = TRUE)

model4_IRxYear_post = as_draws_df(model4, variable = "_cIR:cYear", regex = TRUE)

# get slopes
posterior::summarize_draws(model4_IR_post)
posterior::summarize_draws(model4_year_post)
posterior::summarize_draws(model4_IRxYear_post)

model4_IR_slope_post = model4_IR_post[, 1:8]
colnames(model4_IR_slope_post) = c("Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

model4_year_slope_post = model4_year_post[, 1:8]
colnames(model4_year_slope_post) = c("Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

model4_IRxYear_slope_post = model4_IRxYear_post[, 1:8]
colnames(model4_IRxYear_slope_post) = c("Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

#--- plot slope posteriors ---#

# IR
model4_IR_slope_tidy = pivot_longer(model4_IR_slope_post,
                                    cols = everything(),
                                    names_to = "Mortality")

model4_IR_ridges = ggplot(model4_IR_slope_tidy,
                          aes(x = value, y = Mortality,
                              fill = factor(stat(quantile)))) +
                   theme_bw() +
                   theme(text = element_text(size = 12,  family = "sans")) +
                   stat_density_ridges(geom = "density_ridges_gradient",
                                       calc_ecdf = TRUE,
                                       quantiles = c(0.025, 0.975)) +
                   scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                     guide = "none") +
                   geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                   labs(title = "Model 4: mortality ~ cIR + cYear + cIR:cYear",
                        x = "log odds (IR) relative to 'anthropogenic causes'",
                        y = "Cause of mortality")
model4_IR_ridges

# year
model4_year_slope_tidy = pivot_longer(model4_year_slope_post,
                                      cols = everything(),
                                      names_to = "Mortality")

model4_year_ridges = ggplot(model4_year_slope_tidy,
                            aes(x = value, y = Mortality,
                                fill = factor(stat(quantile)))) +
                     theme_bw() +
                     theme(text = element_text(size = 12,  family = "sans")) +
                     stat_density_ridges(geom = "density_ridges_gradient", calc_ecdf = TRUE,
                             quantiles = c(0.025, 0.975)) +                      
                     scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"),0.6),
                                       guide = "none") +
                     geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                     labs(title = "Model 4: mortality ~ cIR + cYear + cIR:cYear",
                          x = "log odds (year) relative to 'anthropogenic causes'",
                          y = "Cause of mortality")
model4_year_ridges

# interaction, IR x year
model4_IRxYear_slope_tidy = pivot_longer(model4_IRxYear_slope_post,
                                         cols = everything(),
                                         names_to = "Mortality")

model4_IRxYear_ridges = ggplot(model4_IRxYear_slope_tidy,
                               aes(x = value, y = Mortality,
                                   fill = factor(stat(quantile)))) +
                        theme_bw() +
                        theme(text = element_text(size = 17,  family = "sans")) +
                        stat_density_ridges(geom = "density_ridges_gradient",
                                            calc_ecdf = TRUE,
                                            quantiles = c(0.025, 0.975)) +                      
                        scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                          guide = "none") +
                        geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                        labs(#title = "Model 4: mortality ~ cIR + cYear + cIR:cYear",
                             x = "Interaction effect of IR and year
on mortality (log odds)",
                             y = "Cause of mortality") +
                        scale_y_discrete(expand = expansion(add = c(0.5, 2)))
model4_IRxYear_ridges

# save plots 
ggsave(plot = model4_IR_ridges, "plots/bayesian_models/model4_IR.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model4_year_ridges, "plots/bayesian_models/model4_year.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model4_IRxYear_ridges, "plots/bayesian_models/model4_IRxYear.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model4_IRxYear_ridges, "plots/bayesian_models/Figure_9_180mm.pdf", "pdf", width = 180, height = 120, units = "mm")

# resize plot to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/Figure_9_180mm.pdf"
outfile <- "../figures/Figure_9.pdf"

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

### Model 5: Effect of the INTERACTION between IR and age on cause of mortality

Does age influence the effect of inbreeding on cause of mortality?

#### IR x age with all data

```{r model5 - IR x age w/ all data}
model5 = brm(formula = mortality ~ 1 + cIR + cAge + cIR:cAge,
             data = model_data,
             family = categorical(link = logit))
# there are issues with the model, as maximum treedepth is exceeded; per mc-stan.org, this is not as serious as other warnings
# but there are also issues with the tail ESS being too low

# evaluate model performance 
summary(model5)

# posterior predictive check
pp_check(model5, ndraws = 100, type = "bars")

#--- evaluate results ---#
posterior_summary(model5, probs = c(0.055, 0.945))

# IR
mcmc_intervals(model5, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

mcmc_areas(model5, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

mcmc_areas_ridges(model5, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

# age
mcmc_intervals(model5, regex_pars = "_cAge")
mcmc_areas(model5, regex_pars = "_cAge")
mcmc_areas_ridges(model5, regex_pars = "_cAge")

# interaction, IR x age
mcmc_intervals(model5, regex_pars = "_cIR:cAge")
mcmc_areas(model5, regex_pars = "_cIR:cAge")
mcmc_areas_ridges(model5, regex_pars = "_cIR:cAge")

# extract the data
model5_IR_post = as_draws_df(model5, variable = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

model5_age_post = as_draws_df(model5, variable = "_cAge", regex = TRUE)

model5_IRxAge_post = as_draws_df(model5, variable = "_cIR:cAge", regex = TRUE)

# get slopes
posterior::summarize_draws(model5_IR_post)
posterior::summarize_draws(model5_age_post)
posterior::summarize_draws(model5_IRxAge_post)

model5_IR_slope_post = model5_IR_post[, 1:8]
colnames(model5_IR_slope_post) = c("Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

model5_age_slope_post = model5_age_post[, 1:8]
colnames(model5_age_slope_post) = c("Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

model5_IRxAge_slope_post = model5_IRxAge_post[, 1:8]
colnames(model5_IRxAge_slope_post) = c("Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

#--- plot slope posteriors ---#

# IR
model5_IR_slope_tidy = pivot_longer(model5_IR_slope_post,
                                    cols = everything(),
                                    names_to = "Mortality")

model5_IR_ridges = ggplot(model5_IR_slope_tidy,
                          aes(x = value, y = Mortality,
                              fill = factor(stat(quantile)))) +
                   theme_bw() +
                   theme(text = element_text(size = 12,  family = "sans")) +
                   stat_density_ridges(geom = "density_ridges_gradient",
                                       calc_ecdf = TRUE,
                                       quantiles = c(0.025, 0.975)) +
                   scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                     guide = "none") +
                   geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                   labs(title = "Model 5: mortality ~ cIR + cAge + cIR:cAge (all data)",
                        x = "log odds (IR) relative to 'anthropogenic causes'",
                        y = "Cause of mortality")
model5_IR_ridges

# age
model5_age_slope_tidy = pivot_longer(model5_age_slope_post,
                                      cols = everything(),
                                      names_to = "Mortality")

model5_age_ridges = ggplot(model5_age_slope_tidy,
                            aes(x = value, y = Mortality,
                                fill = factor(stat(quantile)))) +
                     theme_bw() +
                     theme(text = element_text(size = 12,  family = "sans")) +
                     stat_density_ridges(geom = "density_ridges_gradient", calc_ecdf = TRUE,
                             quantiles = c(0.025, 0.975)) +                      
                     scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"),0.6),
                                       guide = "none") +
                     geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                     labs(title = "Model 5: mortality ~ cIR + cAge + cIR:cAge (all data)",
                          x = "log odds (age) relative to 'anthropogenic causes'",
                          y = "Cause of mortality")
model5_age_ridges

# interaction, IR x age
model5_IRxAge_slope_tidy = pivot_longer(model5_IRxAge_slope_post,
                                         cols = everything(),
                                         names_to = "Mortality")

model5_IRxAge_ridges = ggplot(model5_IRxAge_slope_tidy,
                               aes(x = value, y = Mortality,
                                   fill = factor(stat(quantile)))) +
                        theme_bw() +
                        theme(text = element_text(size = 12,  family = "sans")) +
                        stat_density_ridges(geom = "density_ridges_gradient",
                                            calc_ecdf = TRUE,
                                            quantiles = c(0.025, 0.975)) +                      
                        scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                          guide = "none") +
                        geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                        labs(title = "Model 5: mortality ~ cIR + cAge + cIR:cAge (all data)",
                             x = "log odds (IRxAge) relative to 'anthropogenic causes'",
                             y = "Cause of mortality")
model5_IRxAge_ridges

# save plots 
ggsave(plot = model5_IR_ridges, "plots/bayesian_models/model5_IR.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model5_age_ridges, "plots/bayesian_models/model5_age.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model5_IRxAge_ridges, "plots/bayesian_models/model5_IRxAge.tiff", "tiff", width = 6, height = 4, units = "in")
```

#### IR x age without neonatal mortality
```{r model5.1 - IR x age w/o neonatal mortality (Figure 7)}
# model
model5.1 = brm(formula = mortality ~ 1 + cIR + cAge + cIR:cAge,
             data = (model_data %>% filter(mortality != "Neonatal mortality")),
             family = categorical(link = logit))

# evaluate model performance 
summary(model5.1)

# posterior predictive check
pp_check(model5.1, ndraws = 100, type = "bars")

#--- evaluate results ---#
posterior_summary(model5.1, probs = c(0.055, 0.945))

# IR
mcmc_intervals(model5.1, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

mcmc_areas(model5.1, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

mcmc_areas_ridges(model5.1, pars = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

# age
mcmc_intervals(model5.1, regex_pars = "_cAge")
mcmc_areas(model5.1, regex_pars = "_cAge")
mcmc_areas_ridges(model5.1, regex_pars = "_cAge")

# interaction, IR x age
mcmc_intervals(model5.1, regex_pars = "_cIR:cAge")
mcmc_areas(model5.1, regex_pars = "_cIR:cAge")
mcmc_areas_ridges(model5.1, regex_pars = "_cIR:cAge")

# extract the data
model5.1_IR_post = as_draws_df(model5.1, variable = c("b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

model5.1_age_post = as_draws_df(model5.1, variable = "_cAge", regex = TRUE)

model5.1_IRxAge_post = as_draws_df(model5.1, variable = "_cIR:cAge", regex = TRUE)

# get slopes
posterior::summarize_draws(model5.1_IR_post)
posterior::summarize_draws(model5.1_age_post)
posterior::summarize_draws(model5.1_IRxAge_post)

model5.1_IR_slope_post = model5.1_IR_post[, 1:7]
colnames(model5.1_IR_slope_post) = c("Bacterial infection", "Dystocia", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

model5.1_age_slope_post = model5.1_age_post[, 1:7]
colnames(model5.1_age_slope_post) = c("Bacterial infection", "Dystocia", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

model5.1_IRxAge_slope_post = model5.1_IRxAge_post[, 1:7]
colnames(model5.1_IRxAge_slope_post) = c("Bacterial infection", "Dystocia", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

#--- plot slope posteriors ---#

# IR
model5.1_IR_slope_tidy = pivot_longer(model5.1_IR_slope_post,
                                    cols = everything(),
                                    names_to = "Mortality")

model5.1_IR_ridges = ggplot(model5.1_IR_slope_tidy,
                          aes(x = value, y = Mortality,
                              fill = factor(stat(quantile)))) +
                   theme_bw() +
                   theme(text = element_text(size = 12,  family = "sans")) +
                   stat_density_ridges(geom = "density_ridges_gradient",
                                       calc_ecdf = TRUE,
                                       quantiles = c(0.025, 0.975)) +
                   scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                     guide = "none") +
                   geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                   labs(title = "Model 5.1: mortality ~ cIR + cAge + cIR:cAge \n(w/o neonatal mortality)",
                        x = "log odds (IR) relative to 'anthropogenic causes'",
                        y = "Cause of mortality")
model5.1_IR_ridges

# age
model5.1_age_slope_tidy = pivot_longer(model5.1_age_slope_post,
                                      cols = everything(),
                                      names_to = "Mortality")

model5.1_age_ridges = ggplot(model5.1_age_slope_tidy,
                            aes(x = value, y = Mortality,
                                fill = factor(stat(quantile)))) +
                     theme_bw() +
                     theme(text = element_text(size = 12,  family = "sans")) +
                     stat_density_ridges(geom = "density_ridges_gradient", calc_ecdf = TRUE,
                             quantiles = c(0.025, 0.975)) +                      
                     scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"),0.6),
                                       guide = "none") +
                     geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                     labs(title = "Model 5.1: mortality ~ cIR + cAge + cIR:cAge \n(w/o neonatal mortality)",
                          x = "log odds (age) relative to 'anthropogenic causes'",
                          y = "Cause of mortality")
model5.1_age_ridges

# interaction, IR x age
model5.1_IRxAge_slope_tidy = pivot_longer(model5.1_IRxAge_slope_post,
                                         cols = everything(),
                                         names_to = "Mortality")

model5.1_IRxAge_ridges = ggplot(model5.1_IRxAge_slope_tidy,
                               aes(x = value, y = Mortality,
                                   fill = factor(stat(quantile)))) +
                        theme_bw() +
                        theme(text = element_text(size = 17,  family = "sans")) +
                        stat_density_ridges(geom = "density_ridges_gradient",
                                            calc_ecdf = TRUE,
                                            quantiles = c(0.025, 0.975)) +                      
                        scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                          guide = "none") +
                        geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                        labs(#title = "Model 5.1: mortality ~ cIR + cAge + cIR:cAge \n(w/o neonatal mortality)",
                             x = "Interaction effect of IR and age
on mortality (log odds)",
                             y = "Cause of mortality") +
                        scale_y_discrete(expand = expansion(add = c(0.5, 2)))
model5.1_IRxAge_ridges

# save plots 
ggsave(plot = model5.1_IR_ridges, "plots/bayesian_models/model5.1_IR_noneonatal.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model5.1_age_ridges, "plots/bayesian_models/model5.1_age_noneonatal.tiff", "tiff", width = 6, height = 4, units = "in")
ggsave(plot = model5.1_IRxAge_ridges, "plots/bayesian_models/Figure_7_180mm.pdf", "pdf", width = 180, height = 120, units = "mm")

# resize plot to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/Figure_7_180mm.pdf"
outfile <- "../figures/Figure_7.pdf"

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

### Model comparison
```{r loo model comparison}
# models with neonatal mortality data
loo(model1, model2, model3, model4, model5, moment_match = TRUE) 

# compare the models based on data w/o neonatal mortality
loo(model3.1, model5.1, moment_match = TRUE)
```