---
title: "Beluga ddRADseq Bayesian models with live biopsy as reference category"
author: "Audrey Salinger"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(brms)
require(posterior)
require(bayesplot)
require(ggplot2)
require(ggridges)
require(tidyr)
require(dplyr)
require(stringr)
require(cowplot)
```

## Objectives

Using Bayesian methods to answer the same questions, this time setting live biopsy category as control rather than trauma, since per Larrat et al. (2024), in contrast to what has been observed in other cetacean populations, SLE belugas dead from trauma do not appear to be in optimal body condition and therefore may not be otherwise healthy as originally presumed.  

**run after primary script**

## Organize the data for Bayesian models

```{r import data}
combined_data = read.delim("combined-data.tsv", sep = "\t")
```

I want to pull out the year from the "date found" data column, which does not have consistent formatting.
```{r extract year found dead}
# create a new column with just year found
combined_data$year_found = str_extract(combined_data$date_found, "\\b\\d{4}\\b")

# check the parsing
combined_data %>% select(c(date_found, year_found))
```
  
I will remove the "degenerative changes" mortality category because it contains only one individual.
```{r data filter}
model_data = combined_data %>% filter(split_inf_category != "Degenerative changes")
```

I will make the nominal predicted variable (cause of death) a factor.  
I will center the metric predictors (age, year found, and inbreeding coefficient) around zero. 
```{r data preparation}
# nominal predicted variable is cause of mortality
# make it a factor and order to put undetermined last
model_data$mortality = factor(model_data$split_inf_category,
                              levels = c("Live biopsy",
                                         "Anthropogenic causes",
                                         "Bacterial infection",
                                         "Dystocia",
                                         "Neonatal mortality",
                                         "Neoplasia",
                                         "Other infection",
                                         "Protozoan infection",
                                         "Verminous infection",
                                         "Undetermined"))

#--- prepare and standardize metric predictor variables ---#

# age
model_data$cAge = model_data$estimated_age - mean(model_data$estimated_age)

# year found
model_data$cYear = as.numeric(model_data$year_found) - mean(as.numeric(model_data$year_found))

# inbreeding coefficients (I will standardize all coeffcients, but I am focusing on running models with IR)
model_data$cIR = model_data$IR_0.001 - mean(model_data$IR_0.001)
model_data$csMLH = model_data$sMLH_0.001 - mean(model_data$sMLH_0.001)
model_data$cF = model_data$F_0.001 - mean(model_data$F_0.001)
model_data$cHL = model_data$HL_0.001 - mean(model_data$HL_0.001)
```

## Model 1: Effect of internal relatedness (IR)

```{r live biopsy - IR}
model1 = brm(formula = mortality ~ 1 + cIR, 
               data = model_data, 
               family = categorical(link = logit))

# evaluate model performance
summary(model1)
mcmc_trace(model1)

# posterior predictive check
pp_check(model1, ndraws = 100, type = "bars")

# evaluate results
mcmc_intervals(model1, regex_pars = "_cIR")
mcmc_areas(model1, regex_pars = "_cIR")
mcmc_areas_ridges(model1, regex_pars = "_cIR")

posterior_summary(model1, probs = c(0.055, 0.945))

# extract the data
model1_post = as_draws_df(model1, variable = "_cIR", regex = TRUE)
posterior::summarize_draws(model1_post)

model1_slope_post = model1_post[, 1:9]
colnames(model1_slope_post) = c("Anthropogenic causes", "Bacterial infection", "Dystocia",
                                "Neonatal mortality", "Neoplasia",
                                "Other infection", "Protozoan infection",
                                "Verminous infection", "Undetermined")

# plot slope posteriors
model1_slope_tidy = pivot_longer(model1_slope_post,
                                 cols = everything(),
                                 names_to = "Mortality")

biopsy_IR_ridges = ggplot(model1_slope_tidy,
                          aes(x = value, y = Mortality,
                              fill = factor(stat(quantile)))) +
                   theme_bw() +
                   theme(text = element_text(size = 17,  family = "sans")) +
                   stat_density_ridges(geom = "density_ridges_gradient",
                                       calc_ecdf = TRUE,
                                       quantiles = c(0.025, 0.975)) +
                   scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                     guide = "none") +
                   geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                   labs(x = "Effect of IR on mortality cause
(log odds, relative to live biopsy)",
                        y = "Cause of mortality") +
                   scale_y_discrete(expand = expansion(add = c(0.5, 2)))
biopsy_IR_ridges

# save plot
ggsave(plot = biopsy_IR_ridges, "plots/bayesian_models/biopsy_IR_180mm.pdf", "pdf", width = 180, height = 120, units = "mm")

# resize plot to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/biopsy_IR_180mm.pdf"
outfile <- "../supplementary_figures/Figure_S4a.pdf"

target_width_mm  <- 80
target_height_mm <- 54

mm_to_pt <- 2.83465
w_pt <- target_width_mm * mm_to_pt
h_pt <- target_height_mm * mm_to_pt

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

## Model 2: Effect of year

```{r live biopsy - year}
model2 =  brm(formula = mortality ~ 1 + cYear, 
              data = model_data, 
              family = categorical(link = logit))

# evaluate model performance
summary(model2)
mcmc_trace(model2)

# posterior predictive check
pp_check(model2, ndraws = 100, type = "bars")

# evaluate results
mcmc_intervals(model2, regex_pars = "_cYear")
mcmc_areas(model2, regex_pars = "_cYear")
mcmc_areas_ridges(model2, regex_pars = "_cYear")

posterior_summary(model2, probs = c(0.055, 0.945))

# extract the data
model2_post = as_draws_df(model2, variable = "_cYear", regex = TRUE)
posterior::summarize_draws(model2_post)

model2_slope_post = model2_post[, 1:9]
colnames(model2_slope_post) = c("Anthropogenic causes", "Bacterial infection", "Dystocia",
                                "Neonatal mortality", "Neoplasia",
                                "Other infection", "Protozoan infection",
                                "Verminous infection", "Undetermined")

# plot slope posteriors
model2_slope_tidy = pivot_longer(model2_slope_post,
                                 cols = everything(),
                                 names_to = "Mortality")

biopsy_year_ridges = ggplot(model2_slope_tidy,
                            aes(x = value, y = Mortality,
                                fill = factor(stat(quantile)))) +
                     theme_bw() +
                     theme(text = element_text(size = 17,  family = "sans")) +
                     stat_density_ridges(geom = "density_ridges_gradient",
                                         calc_ecdf = TRUE,
                                         quantiles = c(0.025, 0.975)) +
                     scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                       guide = "none") +
                     geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                     labs(x = "Effect of year on mortality cause
(log odds, relative to live biopsy)",
                         y = "Cause of mortality") +
                     scale_y_discrete(expand = expansion(add = c(0.5, 2)))

biopsy_year_ridges

# save plot
ggsave(plot = biopsy_year_ridges, "plots/bayesian_models/biopsy_year_180mm.pdf", "pdf", width = 180, height = 120, units = "mm")

# resize plot to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/biopsy_year_180mm.pdf"
outfile <- "../supplementary_figures/Figure_S4b.pdf"

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

## Model 3: Effect of the interaction between IR and year found dead

```{r live biopsy - IR x year}
model3 = brm(formula = mortality ~ 1 + cIR + cYear + cIR:cYear, 
             data = model_data, 
             family = categorical(link = logit))

# evaluate model performance 
summary(model3)

# posterior predictive check
pp_check(model3, ndraws = 100, type = "bars")

#--- evaluate results ---#
posterior_summary(model3, probs = c(0.055, 0.945))

# IR
mcmc_intervals(model3, pars = c("b_muAnthropogeniccauses_cIR", "b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

mcmc_areas(model3, pars = c("b_muAnthropogeniccauses_cIR", "b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

mcmc_areas_ridges(model3, pars = c("b_muAnthropogeniccauses_cIR", "b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

# year
mcmc_intervals(model3, regex_pars = "_cYear")
mcmc_areas(model3, regex_pars = "_cYear")
mcmc_areas_ridges(model3, regex_pars = "_cYear")

# interaction, IR x year
mcmc_intervals(model3, regex_pars = "_cIR:cYear")
mcmc_areas(model3, regex_pars = "_cIR:cYear")
mcmc_areas_ridges(model3, regex_pars = "_cIR:cYear")

# extract the data
model3_IR_post = as_draws_df(model3, variable = c("b_muAnthropogeniccauses_cIR", "b_muBacterialinfection_cIR", "b_muDystocia_cIR", "b_muNeonatalmortality_cIR", "b_muNeoplasia_cIR", "b_muOtherinfection_cIR", "b_muProtozoaninfection_cIR", "b_muVerminousinfection_cIR", "b_muUndetermined_cIR"))

model3_year_post = as_draws_df(model3, variable = "_cYear", regex = TRUE)

model3_IRxYear_post = as_draws_df(model3, variable = "_cIR:cYear", regex = TRUE)

# get slopes
posterior::summarize_draws(model3_IR_post)
posterior::summarize_draws(model3_year_post)
posterior::summarize_draws(model3_IRxYear_post)

model3_IR_slope_post = model3_IR_post[, 1:9]
colnames(model3_IR_slope_post) = c("Anthropogenic causes", "Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

model3_year_slope_post = model3_year_post[, 1:9]
colnames(model3_year_slope_post) = c("Anthropogenic causes", "Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

model3_IRxYear_slope_post = model3_IRxYear_post[, 1:9]
colnames(model3_IRxYear_slope_post) = c("Anthropogenic causes", "Bacterial infection", "Dystocia", "Neonatal mortality", "Neoplasia", "Other infection", "Protozoan infection", "Verminous infection", "Undetermined")

#--- plot slope posteriors ---#
# interaction, IR x year
model3_IRxYear_slope_tidy = pivot_longer(model3_IRxYear_slope_post,
                                         cols = everything(),
                                         names_to = "Mortality")

biopsy_IRxYear_ridges = ggplot(model3_IRxYear_slope_tidy,
                               aes(x = value, y = Mortality,
                                   fill = factor(stat(quantile)))) +
                        theme_bw() +
                        theme(text = element_text(size = 17,  family = "sans"),
                              axis.title.x = element_text(size = 15)) +
                        stat_density_ridges(geom = "density_ridges_gradient",
                                            calc_ecdf = TRUE,
                                            quantiles = c(0.025, 0.975)) +
                        scale_fill_manual(values = alpha(c("#88A2B9", "#657E9D", "#88A2B9"), 0.6),
                                          guide = "none") +
                        geom_vline(xintercept = 0, alpha = 0.5, linetype = "dashed") + 
                        labs(x = "Interaction effect of IR and year on mortality cause
(log odds, relative to live biopsy)",
                            y = "Cause of mortality") +
                        scale_y_discrete(expand = expansion(add = c(0.5, 2)))
biopsy_IRxYear_ridges

# save plot
ggsave(plot = biopsy_IRxYear_ridges, "plots/bayesian_models/biopsy_IRxYear_180mm.pdf", "pdf", width = 180, height = 120, units = "mm")

# resize plot to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/biopsy_IRxYear_180mm.pdf"
outfile <- "../supplementary_figures/Figure_S4c.pdf"

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

Arrange plots in a grid.
```{r cowplot grid}
biopsy_model_grid <-  plot_grid(biopsy_IR_ridges, 
                                biopsy_year_ridges,
                                biopsy_IRxYear_ridges,
                      nrow = 3, ncol = 1,
                      labels = c("(a)","(b)","(c)","(d)"), label_size = 20, hjust = -0.6, vjust = 2)
biopsy_model_grid

# save grid
ggsave(plot = biopsy_model_grid, "plots/bayesian_models/biopsy_model_grid_180mm.pdf", "pdf", width = 180, height = 360, units = "mm")

# resize grid to 80 mm width using GhostScript
infile  <- "plots/bayesian_models/biopsy_model_grid_180mm.pdf"
outfile <- "../supplementary_figures/Figure_S4.pdf"

target_width_mm  <- 80
target_height_mm <- 54*3

mm_to_pt <- 2.83465
w_pt <- target_width_mm * mm_to_pt
h_pt <- target_height_mm * mm_to_pt

cmd <- sprintf("gs -o \"%s\" -sDEVICE=pdfwrite -dDEVICEWIDTHPOINTS=%.2f -dDEVICEHEIGHTPOINTS=%.2f -dPDFFitPage \"%s\"",
               outfile, w_pt, h_pt, infile)
system(cmd)
```

